---
title: "MolBreeding 45K array vs SNP50K skim — ancestry-call concordance"
author: "Fausto Rodríguez-Zapata"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    df-print: paged
    embed-resources: true
execute:
  warning: false
  message: false
params:
  mb_calls:   "data/molbreeding_45k/calls_common_schema.csv"
  skim_calls: "data/rtiger_50K/calls_common_schema.csv"
  sample_map: "data/molbreeding_45k_sample_map.tsv"
  mb_counts:  "data/molbreeding_45k/counts"
---

**Repo:** [sawers-rellan-labs/zealtiger](https://github.com/sawers-rellan-labs/zealtiger)

Cross-validation of ancestry source #2 (SNP50K skim → RTIGER, ~0.4×) against source
#5 (MolBreeding 45K SNP array, ~100% call) for the 14 BZea NILs genotyped on both
platforms. The array is the higher-confidence empirical genotyping; the skim is the
low-coverage workhorse used genome-wide. Both call sets are RTIGER segments on B73 NAM
v5 in the common schema (`name, chr, start_bp, end_bp, state` with 0 = homozygous
recurrent / B73, 1 = het, 2 = homozygous donor / teosinte).

The consensus-map BC2S3 **simulation** (source #1) is shown only as a **null
expectation** for introgression structure under the modeled recombination process
(MaizeGDB consensus map, ~1,783 cM — inflated for a composite map; `m=10` interference;
BC2S3 pedigree). It is **not ground truth** — deviations from it may reflect the model's
biases as much as the data's.

> Provenance of the inputs (run first): the MolBreeding panel is lifted v3→v4→v5
> (`molbreeding_liftover.R`), encoded as 0/1/2 pseudo-counts (`molbreeding_to_rtiger.R`),
> and fit at rigidity r = 2 (`make_45k_array_benchmark.R` + `sweep_rigidity_45k.R`;
> the skim uses r = 3). The skim per-donor calls come from `fit_rtiger_by_donor.R`.

## Setup

**Purpose**: Load the two call sets and the sample map; define the evaluation grid.

**Approach**: The grid is the MolBreeding marker set (the ~21k informative v5
positions the array actually measured) — the fairest common ground. Each source's
state at each grid point is read by segment overlap.

```{r setup}
library(tidyverse)
library(GenomicRanges)
library(patchwork)

mb   <- read_csv(params$mb_calls,   show_col_types = FALSE)
skim <- read_csv(params$skim_calls, show_col_types = FALSE)
smap <- read_tsv(params$sample_map, show_col_types = FALSE)

# evaluation grid = MolBreeding 45K marker positions (v5)
cf <- read_tsv(list.files(params$mb_counts, full.names = TRUE)[1],
               col_names = c("chr_label", "bp", "rb", "rc", "ab", "ac"),
               show_col_types = FALSE)
grid <- tibble(chr = as.integer(sub("chr", "", cf$chr_label)), bp = cf$bp)

samples <- intersect(unique(mb$name), unique(skim$name))   # 14 NILs on both platforms
cat(sprintf("Samples on both platforms: %d | grid markers: %d\n",
            length(samples), nrow(grid)))
```

## State on the common grid

**Purpose**: Assign each source a 0/1/2 state at every grid marker, per sample.

**Approach**: Build per-sample GRanges of segments; overlap the width-1 grid
positions to pull the called state.

```{r state-grid}
state_at <- function(df, s, grid) {
  d <- df[df$name == s, ]
  g <- GRanges(d$chr, IRanges(d$start_bp, d$end_bp))
  q <- GRanges(grid$chr, IRanges(grid$bp, width = 1))
  d$state[findOverlaps(q, g, select = "first")]
}

paired <- map_dfr(samples, function(s) {
  a <- state_at(mb, s, grid); k <- state_at(skim, s, grid)
  ok <- !is.na(a) & !is.na(k)
  tibble(name = s, array = a[ok], skim = k[ok])
})
paired <- paired %>%
  left_join(smap %>% transmute(name = sample_id, accession = sub("_.*$", "", genotype)),
            by = "name")
```

## Concordance vs skim missingness

**Purpose**: Test whether the skim's agreement with the array degrades as the skim's
own missing-data rate rises — i.e. is the disagreement coverage-driven?

**Approach**: Per-sample skim missingness = fraction of the SNP50K sites with no
reads (`RefCount + AltCount == 0`), read from the skim input counts
(`data/rtiger_50K/counts/<accession>/<sample>.tsv`). Plot against the per-sample
array-vs-skim concordance.

```{r per-sample}
skim_missing <- function(s) {
  f <- Sys.glob(file.path("data/rtiger_50K/counts", "*", paste0(s, ".tsv")))
  if (!length(f)) return(NA_real_)
  d <- readr::read_tsv(f[1], col_names = c("chr","pos","rb","rc","ab","ac"),
                       show_col_types = FALSE)
  mean(d$rc + d$ac == 0)
}

per_sample <- paired %>%
  group_by(name, accession) %>%
  summarise(n = n(), concordance = mean(array == skim),
            array_dosage = mean(array) / 2, .groups = "drop") %>%
  mutate(skim_missing = map_dbl(name, skim_missing)) %>%
  arrange(desc(skim_missing))
per_sample
```

```{r conc-vs-missing}
r_cm <- cor(per_sample$skim_missing, per_sample$concordance)
ggplot(per_sample, aes(skim_missing, concordance)) +
  geom_smooth(method = "lm", se = FALSE, colour = "grey75", linewidth = 0.5) +
  geom_point(aes(colour = array_dosage), size = 3) +
  geom_text(aes(label = name), size = 2.6, vjust = -0.9, check_overlap = TRUE) +
  scale_x_continuous(labels = scales::percent) +
  scale_colour_viridis_c(option = "magma", direction = -1, end = 0.9) +
  labs(x = "skim missing-data rate (fraction of SNP50K sites uncovered)",
       y = "marker concordance with 45K array",
       colour = "donor\ndosage\n(array)",
       title = sprintf("Concordance vs skim missingness (r = %+.2f)", r_cm),
       subtitle = "Discordance tracks donor content, not missingness — the low-concordance sample is the most donor-rich") +
  theme_bw()
```

## Confusion matrix — array (reference) vs skim

Rows = array state (reference, MolBreeding), columns = skim call; each cell is the marker count and
the row percentage (the skim's recovery of that array state). The diagonal is
correct recovery; off-diagonal shows where the skim drifts.

```{r confusion}
cm      <- table(array = factor(paired$array, 0:2), skim = factor(paired$skim, 0:2))
overall <- sum(diag(cm)) / sum(cm)
prop    <- sweep(cm, 1, rowSums(cm), "/")
disp    <- matrix(sprintf("%d (%.1f%%)", cm, 100 * prop), nrow = 3,
                  dimnames = list(`array \\ skim` = c("0 recurrent","1 het","2 donor"),
                                  c("0 recurrent","1 het","2 donor")))
knitr::kable(disp, align = "r",
             caption = sprintf("Array (rows) vs skim (cols): count (row %%). Overall marker concordance %.3f.",
                               overall))

# donor-present recovery
dp <- paired$array %in% c(1, 2)
cat(sprintf("Array donor-present markers: %d | skim also donor-present: %.1f%%\n",
            sum(dp), 100 * mean(paired$skim[dp] %in% c(1, 2))))
cat(sprintf("Array homozygous-donor markers: %d | skim also hom-donor: %.1f%%\n",
            sum(paired$array == 2), 100 * mean(paired$skim[paired$array == 2] == 2)))
```

## Donor-call composition — het vs homozygous donor

**Purpose**: With the dominant homozygous-recurrent (B73) background removed, zoom in
on the donor calls and compare their composition between platforms.

**Approach**: For each sample × platform, stack het and homozygous-donor as their
percentage of the (shared, covered) markers — so the bar height is the true donor
fraction of the genome (a few % up to ~16%), with the dominant homozygous-recurrent
background simply not drawn. Array next to skim, per sample (ordered by array donor
dosage). The skim's shorter, teal-shifted bars show donor under-calling and the
homozygous-donor → het demotion.

```{r donor-composition}
denom <- paired %>% count(name, name = "tot")     # shared markers covered on both platforms
comp <- bind_rows(
  paired %>% transmute(name, platform = "array (45K)", state = array),
  paired %>% transmute(name, platform = "skim (50K)",  state = skim)
) %>%
  filter(state %in% c(1, 2)) %>%
  count(name, platform, state) %>%
  left_join(denom, by = "name") %>%
  mutate(pct = n / tot, state = factor(state, c(1, 2), c("het", "homozygous donor")))

lev <- per_sample %>% arrange(array_dosage) %>% pull(name)
comp$name <- factor(comp$name, levels = lev)

ggplot(comp, aes(y = name, x = pct, fill = state)) +
  geom_col(position = "stack") +
  facet_wrap(~ platform) +
  scale_fill_manual(values = c("het" = "forestgreen", "homozygous donor" = "purple4")) +
  scale_x_continuous(labels = scales::percent) +
  labs(x = "% of genome called donor (het + homozygous donor)", y = NULL, fill = NULL,
       title = "Donor calls as % of genome (recurrent B73 not drawn)",
       subtitle = "Stacked het + homozygous donor; array (reference) vs skim, per sample") +
  theme_bw()
```

## Donor dosage — array vs skim

**Approach**: Per-sample donor dosage on the common grid = mean(state)/2. Includes
the 2 B73-bulk checks (expected ≈ 0) as negative controls.

```{r dosage}
dose <- bind_rows(
  map_dfr(union(samples, c("PN4_SID326", "PN6_SID576")), function(s) {
    a <- state_at(mb, s, grid)
    tibble(name = s, source = "array", dosage = mean(a[!is.na(a)]) / 2)
  }),
  map_dfr(samples, function(s) {
    k <- state_at(skim, s, grid)
    tibble(name = s, source = "skim", dosage = mean(k[!is.na(k)]) / 2)
  })
) %>%
  pivot_wider(names_from = source, values_from = dosage) %>%
  left_join(smap %>% transmute(name = sample_id, genotype), by = "name") %>%
  mutate(kind = if_else(grepl("^B73", genotype), "B73 check", "teosinte NIL"))

ggplot(dose, aes(array, skim, colour = kind)) +
  geom_abline(slope = 1, linetype = 2, colour = "grey50") +
  geom_point(size = 2) +
  scale_colour_manual(values = c("B73 check" = "grey50", "teosinte NIL" = "purple4")) +
  coord_equal() +
  labs(x = "donor dosage — array (MolBreeding 45K)",
       y = "donor dosage — skim (SNP50K)",
       colour = NULL,
       title = "Skim systematically under-calls donor dosage vs the array") +
  theme_bw()
```

## Introgression-size distribution

**Purpose**: Compare the sizes of called donor introgressions between the array and
the skim — does the low-coverage skim fragment introgressions into smaller pieces?

**Approach**: An introgression = a maximal run of consecutive donor-present segments
(state 1 or 2) per chromosome; its size is the span end−start. Sizes pooled across the
14 shared NILs, array (forestgreen) vs skim (purple4).

```{r intro-sizes}
intro_sizes <- function(df, plat) {
  df %>%
    filter(name %in% samples) %>%
    arrange(name, chr, start_bp) %>%
    group_by(name, chr) %>%
    mutate(donor = state > 0, run = cumsum(c(TRUE, diff(donor) != 0))) %>%
    filter(donor) %>%
    group_by(name, chr, run) %>%
    summarise(size_mb = (max(end_bp) - min(start_bp)) / 1e6, .groups = "drop") %>%
    mutate(platform = plat)
}
sizes <- bind_rows(intro_sizes(mb, "array (45K)"),
                   intro_sizes(skim, "skim (50K)"))

# consensus-map BC2S3 simulation (null expectation), from the 45K-array benchmark mosaic
sim_seg <- readr::read_csv("results/rtiger_45k_benchmark/truth_segments.csv",
                             show_col_types = FALSE)
sim_sizes <- sim_seg %>%
  arrange(name, chr, start_bp) %>%
  group_by(name, chr) %>%
  mutate(donor = state > 0, run = cumsum(c(TRUE, diff(donor) != 0))) %>%
  filter(donor) %>%
  group_by(name, chr, run) %>%
  summarise(size_mb = (max(end_bp) - min(start_bp)) / 1e6, .groups = "drop")

bind_rows(sizes, mutate(sim_sizes, platform = "simulation (null)")) %>%
  group_by(platform) %>%
  summarise(n_introgressions = n(), median_mb = round(median(size_mb), 2),
            mean_mb = round(mean(size_mb), 2), .groups = "drop")
```

```{r intro-sizes-plot}
ggplot(sizes, aes(size_mb)) +
  geom_histogram(aes(y = after_stat(density), fill = platform),
                 position = "identity", alpha = 0.5, bins = 40) +
  geom_freqpoly(data = sim_sizes, aes(y = after_stat(density)),
                bins = 40, colour = "grey20", linewidth = 0.8) +
  scale_fill_manual(values = c("array (45K)" = "forestgreen", "skim (50K)" = "purple4")) +
  scale_x_log10() +
  labs(x = "introgression size (Mb, log scale)", y = "density", fill = NULL,
       title = "Introgression-size distribution: array vs skim",
       subtitle = "grey line = consensus-map simulation (null expectation)") +
  theme_bw()
```

## Findings

- Mean marker concordance ≈ 0.91 across the 14 shared NILs.
- **Discordance is driven by donor content, not skim missingness.** Concordance vs
  skim missing-rate is weakly positive (r ≈ +0.3) — more dropout does not mean worse
  agreement, because the surviving markers are mostly recurrent background that both
  platforms call. The lowest-concordance sample is the most donor-rich, not the most
  missing: the skim errs specifically where there is donor signal.
- **Recurrent background is recovered near-perfectly** (~96% of array-recurrent
  markers), so the skim is reliable for the B73 genome and for detecting donor
  presence (~71% of donor-present markers recovered).
- **The skim under-calls donor dosage**, and homozygous donor is the weak spot
  (~17% recovered as BB; most demoted to het or recurrent) — at ~0.4× a
  homozygous-donor site rarely shows enough ALT reads to read as BB. This is the
  coverage-limited behaviour anticipated in `docs/12-het-excess-diagnosis.md`, here
  quantified against an empirical reference.
- **Relative to the consensus-map simulation (null expectation, grey).** At r=2 the
  array shows many small segments and the skim few large ones; both are compared to
  the simulation only as a null baseline, not ground truth. Segment granularity is
  strongly rigidity-dependent (see the separate rigidity analysis): at a matched ~2 Mb
  resolution the array resolves ~13 donor blocks/sample and the skim ~6, vs the
  simulation's ~10 — so the empirical platforms bracket the null rather than "err" against it.
- The B73-bulk checks sit at ≈ 0 dosage on both platforms (polarity / negative control
  confirmed).
```
