---
title: "B73 check QC — cross-experiment donor content"
author: "Fausto Rodríguez-Zapata"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    df-print: paged
    embed-resources: true
execute:
  warning: false
  message: false
params:
  brb_calls:   "results/sim_calibration/brbseq_taxa_r7_wideseq/calls_common_schema.csv"  # BrB-RTI, per-species, wideseq-thinned r=7
  brb_map:     "agent/brb_species_map.csv"                                               # BRB name -> accession (B73 flag)
  skim_checks: "data/rtiger_50K/calls_checks_common_schema.csv"                          # Skim-RTI checks (B73, Purple)
  molb_calls:  "results/sim_calibration/gbts_calls/SNP_wsfilt/calls_common_schema.csv"   # MolB-RTI (B73-bulk)
---

**Repo:** [sawers-rellan-labs/zealtiger](https://github.com/sawers-rellan-labs/zealtiger)

A **B73 control** must be a *pure recurrent-parent* rep — **no introgression** (≈ 0 donor
genome-wide). This doc tracks **every B73 rep across the three experiments** (BrB-seq,
SNP50K skim, MolBreeding GBTS), scores each rep's donor content from its own caller, and
flags reps that are **not actually clean B73** (localized donor on a chromosome →
mislabel / contamination / real introgression). It exists because the BrB-seq B73 control
`PN1_SID15` turned out to be an impostor (a chromosome arm reads ~50% donor).

- **donor dosage** = genome-wide donor-allele fraction (`Σ len·state/2 / Σ len`, state REF=0/HET=1/ALT=2); 0 = pure B73.
- **maxchr** = the worst single-chromosome donor fraction — localizes an introgressed arm.
- **flag**: `clean` (dosage < 0.02), `borderline` (0.02–0.10), **`impostor`** (> 0.10).

```{r setup}
library(tidyverse); library(ggpubr)
pal <- c(REF = "gold", HET = "springgreen4", ALT = "purple4")

brbmap <- read_csv(params$brb_map, show_col_types = FALSE)
brb_b73 <- brbmap$name[grepl("^B73", brbmap$accession)]
calls <- bind_rows(
  read_csv(params$brb_calls,  show_col_types = FALSE) %>% filter(name %in% brb_b73) %>%
    mutate(exp = "BrB-RTI (r=7, wideseq)"),
  read_csv(params$skim_checks, show_col_types = FALSE) %>% filter(donor == "B73") %>%
    mutate(exp = "Skim-RTI (r=5)"),
  read_csv(params$molb_calls,  show_col_types = FALSE) %>% filter(donor == "B73-bulk") %>%
    mutate(exp = "MolB-RTI (r=2)")
) %>% select(exp, name, chr, start_bp, end_bp, state)

qc <- calls %>% mutate(l = end_bp - start_bp + 1) %>%
  group_by(exp, name) %>%
  summarise(dosage = sum(l * state / 2) / sum(l),
            maxchr = max(tapply(l * state / 2, chr, sum) / tapply(l, chr, sum)),
            worst_chr = paste0("chr", names(which.max(tapply(l * state / 2, chr, sum) / tapply(l, chr, sum)))),
            .groups = "drop") %>%
  mutate(flag = case_when(dosage > 0.10 ~ "impostor", dosage > 0.02 ~ "borderline", TRUE ~ "clean")) %>%
  arrange(desc(dosage))
cat(sprintf("B73 reps: %d (BrB %d, Skim %d, MolB %d) | clean %d | borderline %d | impostor %d\n",
            nrow(qc), sum(qc$exp == "BrB-RTI (r=7, wideseq)"), sum(qc$exp == "Skim-RTI (r=5)"),
            sum(qc$exp == "MolB-RTI (r=2)"),
            sum(qc$flag == "clean"), sum(qc$flag == "borderline"), sum(qc$flag == "impostor")))
```

## Per-rep donor content (all experiments, worst → best)

```{r table}
qc %>% transmute(experiment = exp, sample = name, dosage = round(dosage, 3),
                 maxchr = round(maxchr, 3), worst_chr, flag) %>%
  knitr::kable(caption = "Every B73 rep, by donor dosage. `impostor` = not pure B73 (a chromosome carries donor).")
```

**Read:** the `impostor` reps carry **localized** donor (high `maxchr` on one chromosome) —
the signature of a real introgressed segment, not diffuse noise — so they are mislabeled /
contaminated samples, not B73. The BrB-seq impostors cluster on plate PN1.

## Chromosome painting — 5 cleanest vs 5 worst B73 reps

```{r painting, fig.width=12, fig.height=7}
ord <- qc %>% arrange(desc(dosage))
sel <- bind_rows(head(ord, 5) %>% mutate(grp = "5 WORST (impostor)"),
                 tail(ord, 5) %>% mutate(grp = "5 BEST (clean)"))
lab <- function(e, n, d) sprintf("%s<br>%s<br>%.3f", sub(" .*", "", e), n, d)
paint <- calls %>% inner_join(sel %>% select(exp, name, dosage, grp), by = c("exp", "name")) %>%
  mutate(row = lab(exp, name, dosage),
         state = factor(state, 0:2, c("REF", "HET", "ALT")),
         chr = factor(chr, levels = 1:10, labels = paste0("chr", 1:10)),
         grp = factor(grp, c("5 WORST (impostor)", "5 BEST (clean)")))
row_ord <- sel %>% mutate(row = lab(exp, name, dosage)) %>% arrange(grp, desc(dosage)) %>% pull(row)
paint$row <- factor(paint$row, levels = rev(row_ord))

ggplot(paint, aes(xmin = start_bp/1e6, xmax = end_bp/1e6, ymin = 0, ymax = 1, fill = state)) +
  geom_rect() +
  facet_grid(row ~ chr, scales = "free_x", space = "free_x", switch = "y") +
  scale_fill_manual(values = pal, drop = FALSE) +
  scale_x_continuous(labels = scales::unit_format(unit = "", scale = 1)) +
  labs(title = "B73 reps — 5 worst (impostors) vs 5 cleanest",
       subtitle = "row = experiment / sample / donor dosage | gold REF, green HET, purple ALT — a clean B73 is all gold",
       x = "Position (Mb)", fill = "Genotype") +
  ggpubr::theme_classic2(base_size = 9) +
  theme(legend.position = "bottom", strip.background = element_blank(),
        strip.text.x = element_text(face = "bold"),
        strip.text.y.left = ggtext::element_markdown(angle = 0, hjust = 1, size = 8, lineheight = 0.85),
        panel.spacing.x = unit(0.05, "lines"), panel.spacing.y = unit(0.15, "lines"),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        axis.title.y = element_blank(), axis.line.y = element_blank(),
        axis.text.x = element_text(size = 6))
```

## Recommended clean controls

```{r controls}
qc %>% filter(flag == "clean") %>% group_by(exp) %>% slice_min(dosage, n = 3, with_ties = FALSE) %>%
  transmute(experiment = exp, sample = name, dosage = round(dosage, 4), maxchr = round(maxchr, 4)) %>%
  knitr::kable(caption = "Cleanest 3 B73 reps per experiment — use one of these as the control (NOT PN1_SID15).")
```

**Use:** for the BrB-RTI comparison docs, replace the `PN1_SID15` control with the cleanest
BrB-seq rep above. The `impostor` reps should be excluded from any B73-check pooling and
flagged for the BRB-seq relabeling follow-up (`docs/brbseq-sample-mislabeling.md`).
