---
title: "Fragment-size distribution in cM — theory and cross-caller validation"
author: "Fausto Rodríguez-Zapata"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    df-print: paged
    embed-resources: true
execute:
  warning: false
  message: false
params:
  skim_calls: "data/rtiger_50K/calls_taxa_r5.csv"                                       # Skim-RTI, per-taxon r=5
  brb_taxa:   "results/sim_calibration/brbseq_taxa_r7_wideseq/calls_common_schema.csv"   # BrB-RTI r=7, wideseq-filtered, per teosinte species
  skimbin:    "results/sim_calibration/skimbin_calls/calls_common_schema.csv"           # Skim-BIN: BzeaSeq 1 Mb-bin GMM-HMM (same skim reads as Skim-RTI)
  nilhmm:     "results/sim_calibration/nilhmm_calls/calls_common_schema.csv"            # nilHMM: count-based BetaBinomial HMM, per-taxon r (same skim reads as Skim-RTI)
  corr:       "results/sample_correspondence/skim_brbseq_correspondence.csv"            # curated NIL pairs (pedigree-matched)
  nil_seg:    "results/nil_segments.csv"                                                # n=1500 BC2S3 simulation segments — carries BOTH cM and Mb endpoints
  map:        "results/maize_map_v5_clean.rds"                                          # consensus Marey map: (chr, bp, cm) anchors, B73 NAM v5
---

**Repo:** [sawers-rellan-labs/zealtiger](https://github.com/sawers-rellan-labs/zealtiger)

The [introgression-size panel](skim_brbseq_skimbin.html) compares the four callers
(Skim-RTI / BrB-RTI / Skim-BIN / nilHMM) against the n=1500 BC2S3 simulation **in megabases**. The
Mb fragment-size law has **no closed form**: it is the genetic-distance (cM) law pushed
through the cM→Mb change of variables $x_{\text{Mb}}=\int dx_{\text{cM}}/\rho(x)$, and the
local recombination rate $\rho(x)=d\text{cM}/d\text{Mb}$ is an *empirical, position-dependent*
lookup (strong pericentromeric suppression, hot arms) — so the simulation is doing Monte-Carlo
integration over the map. **In cM the segment-length law is analytic.** This document

1. **derives** the theoretical donor-segment-size distribution in cM from the BC2S3
   recombination model,
2. confirms the derived law reproduces the simcross output (its own cM segment lengths), and
3. uses it as a **map-free reference** to validate the four callers, after converting their
   called Mb segments back to cM through the consensus map.

> **Why this is a *validation* panel, not the calibration metric.** A caller's resolution
> limit lives in Mb (Skim-BIN's hard 1 Mb floor; RTIGER's marker-space rigidity `r`).
> Converting to cM turns a clean Mb floor into a *position-dependent* cM floor (≈0 cM at a
> centromere, tens of cM on an arm), entangling caller resolution with map non-linearity. So
> the [Mb panel](skim_brbseq_skimbin.html) stays the `r`-calibration space; this cM panel
> tests whether the *population's* fragment law matches the analytic recombination
> expectation, with the map entering **once, deterministically, on the data side**. It leans
> on **conserved marker order** relative to the consensus map — load-bearing here, since any
> introgression with local rearrangement makes $\rho(x)$ wrong and biases $\Delta$cM.

# Part 1 — The theoretical fragment-size distribution in cM

## The crossover process (Haldane)

Under Haldane's no-interference model, crossovers along a chromosome form a **homogeneous
Poisson process in genetic distance**: 1 Morgan = 1 expected crossover, and the count in any
interval of length $\ell$ Morgans is $\text{Poisson}(\ell)$. Consequently the distance to the
next crossover is **exponential**,

$$P(D > x) = e^{-x}, \qquad x \text{ in Morgans},$$

with mean 1 Morgan, $\text{CV}=1$, skewness $=2$ — the most-skewed member of the gamma family
($\text{Gamma}(1)$). Genetic distance is literally the scale parameter: it *is* the expected
number of crossovers, so it plays the role temperature plays for a Maxwell–Boltzmann speed.

## From crossovers to junctions: superposition over meioses

A donor↔recurrent boundary ("junction") on the final NIL haplotype is a crossover — laid down
in *some* meiosis of the NIL's ancestry — that happened to fall where one parental haplotype
carried donor ancestry and the other recurrent. If the NIL haplotype descends through $g$
meioses since the donor haplotype was intact, and each meiosis sprinkles an independent
Poisson($1/$Morgan) crossover field, the **superposition** of those $g$ fields is itself
Poisson at rate $g$ per Morgan. (Exact under Haldane; and by the Palm–Khinchin superposition
theorem, the pooled process tends to Poisson *even with per-meiosis interference*, because
independent point processes superpose toward Poisson.)

If every crossover were a junction, junction density would be $g$/Morgan and donor tracts would
be $\text{Exp}(g)$. The correction is that **only crossovers in still-heterozygous (donor/recurrent)
territory create a D↔R junction**, and that territory shrinks each generation as the donor
fraction is diluted. So the effective junction rate is $\lambda \le g$.

## Counting meioses in the BC2S3 pedigree

The simulated scheme (`R/02_simulate.R`, `bc2s3_pedigree()`) is two backcrosses to the
recurrent parent then three generations of selfing; the donor founder (id 2) is inbred, so the
donor haplotype enters the F1 (id 3) **intact** and is first broken in the F1's meiosis. Tracing
one homolog of the final NIL (id 8) back to that intact donor haplotype crosses **five**
meioses:

| # | meiosis (gametogenesis in) | produces | territory still D/R |
|---|---|---|---|
| 1 | F1 (id 3) | BC1 (id 4) | **whole genome** (F1 het everywhere) |
| 2 | BC1 (id 4) | BC2 (id 5) | donor regions only (~50% of remaining) |
| 3 | BC2 (id 5) | S1  (id 6) | shrinking donor regions |
| 4 | S1  (id 6) | S2  (id 7) | shrinking donor regions |
| 5 | S2  (id 7) | S3  (id 8) = NIL | shrinking donor regions |

So $g = 5$. The naïve Haldane prediction is therefore $\text{Exp}(5)$ per Morgan — donor tracts
with mean $1/5$ Morgan $= 20$ cM. Because junction-creating territory shrinks after meiosis 1,
the *effective* rate is a bit below 5.

## The donor-tract length distribution

Two quantities, distinguished by the **inspection paradox**:

- **Genome-wide tracts** (a randomly chosen donor segment, no selection): by renewal theory the
  length is $\text{Exp}(\lambda)$, mean $1/\lambda$ Morgans.
- **The tract covering a fixed target locus** (a selected NIL around a target gene): *length-biased*
  → left flank + right flank, each $\text{Exp}(\lambda)$, so $T^\star \sim \text{Gamma}(2,\lambda)$,
  mean $2/\lambda$, skewness $\sqrt2\approx1.41$. This is the $\text{Gamma}(2)$ "segment around a
  locus" case.

These simulations are **unselected**, so the **genome-wide** quantity applies — the law sits near
the exponential end, not at $\text{Gamma}(2)$.

## Three departures from the clean exponential

The simulated segments are not exactly $\text{Exp}(5)$ for three reasons, each pushing in a
definite direction:

1. **Union of the two haplotypes.** A simulated "segment" is the *union* of donor intervals
   across the mat and pat homologs (a region is introgressed if *either* carries donor). Unions
   merge nearby tracts and lengthen them → mean up, and the shortest tracts are absorbed → **shape
   $k>1$**.
2. **Interference ($m=10$).** simcross uses the Stahl chi-square model with $m=10,\,p=0$ (a
   crossover at every 11th point of a Poisson process) — strong positive interference, which
   *regularizes* spacing within each meiosis → $\text{CV}<1$, skewness $<2$, again **$k>1$**.
3. **Shrinking heterozygous territory.** Only meiosis 1 acts genome-wide; later meioses add
   junctions only in the dwindling donor neighborhoods → **effective rate $\lambda<g=5$**.

## The theoretical distribution we will test against

Net: the donor-tract length in cM is well-described by a **gamma law**

$$T \sim \text{Gamma}(k,\lambda), \qquad k \gtrsim 1,\ \lambda \lesssim 5\ \text{Morgan}^{-1},$$

**bracketed below by $\text{Exp}(5)$** (no interference, no union, full rate) and **above by
$\text{Gamma}(2)$** (the locus-anchored / heavily size-biased extreme). Part 2 estimates
$(k,\lambda)$ from the simulation and checks both brackets; Part 3 uses the fitted analytic law
plus the simulation's own cM lengths as the reference for the callers.

```{r setup}
library(tidyverse)
theme_set(theme_bw())

nil_seg <- read_csv(params$nil_seg, show_col_types = FALSE)
# simulated segments carry BOTH endpoints; the cM length is exact (no map needed) — this IS the
# generative model's realization of the law derived above.
sim_cm <- nil_seg %>% transmute(cm = end_cm - start_cm, M = (end_cm - start_cm) / 100)

g  <- 5                                   # meioses on the donor lineage (pedigree count)
mu <- mean(sim_cm$cm)                     # empirical mean (cM)

cat(sprintf("sim cM segments: n=%d | mean %.2f  median %.2f  sd %.2f  CV %.3f  skew %.3f\n",
            nrow(sim_cm), mu, median(sim_cm$cm), sd(sim_cm$cm),
            sd(sim_cm$cm)/mu, mean(((sim_cm$cm - mu)/sd(sim_cm$cm))^3)))
```

# Part 2 — Does the derived law reproduce the simulation?

We fit $\text{Gamma}(k,\lambda)$ to the simulation's own cM segment lengths by maximum
likelihood, and overlay the two analytic brackets, $\text{Exp}(5)$ and the mean-matched
$\text{Gamma}(2)$.

```{r fit-gamma}
fit <- MASS::fitdistr(sim_cm$M, "gamma")           # in Morgans
k   <- unname(fit$estimate["shape"])
lam <- unname(fit$estimate["rate"])                # per Morgan
ks_gamma <- suppressWarnings(ks.test(sim_cm$M, "pgamma", shape = k, rate = lam)$statistic)
ks_exp5  <- suppressWarnings(ks.test(sim_cm$M, "pexp",  rate = g)$statistic)
ks_gam2  <- suppressWarnings(ks.test(sim_cm$M, "pgamma", shape = 2, rate = 2/mean(sim_cm$M))$statistic)

cat(sprintf("Gamma MLE (cM): shape k=%.3f  rate lambda=%.3f /Morgan  -> mean %.2f cM, CV %.3f, skew %.3f\n",
            k, lam, 100*k/lam, 1/sqrt(k), 2/sqrt(k)))
cat(sprintf("KS vs derived Gamma(k,lambda): D=%.4f   [Exp(5): D=%.4f | Gamma(2): D=%.4f]\n",
            ks_gamma, ks_exp5, ks_gam2))
```

The fitted **$k\approx`r round(k,2)`$, $\lambda\approx`r round(lam,2)`$/Morgan** lands exactly
where the derivation predicts: just above the exponential ($k>1$, from interference + the union)
with a rate a little under $g=5$ (shrinking het territory). It fits the simulation tightly
(KS $D\approx`r round(ks_gamma,3)`$), while both brackets — $\text{Exp}(5)$ and $\text{Gamma}(2)$
— are rejected, confirming the segment law sits *between* them, near the exponential end.

```{r plot-theory, fig.width=8, fig.height=4.6}
xs <- seq(0, quantile(sim_cm$cm, 0.995), length.out = 400)
curves <- bind_rows(
  tibble(cm = xs, d = dgamma(xs/100, shape = k, rate = lam)/100,
         curve = sprintf("derived Gamma(k=%.2f, λ=%.1f)", k, lam)),
  tibble(cm = xs, d = dexp(xs/100, rate = g)/100,           curve = "Exp(5)  [lower bracket]"),
  tibble(cm = xs, d = dgamma(xs/100, shape = 2, rate = 2/mean(sim_cm$M))/100,
         curve = "Gamma(2)  [upper bracket]"))

ggplot() +
  geom_histogram(data = sim_cm, aes(cm, after_stat(density)), bins = 40,
                 fill = "grey80", colour = "grey55", linewidth = 0.2) +
  geom_line(data = curves, aes(cm, d, colour = curve, linetype = curve), linewidth = 0.9) +
  scale_colour_manual(values = c("black", "firebrick", "royalblue3")) +
  scale_linetype_manual(values = c("solid", "dashed", "dashed")) +
  labs(x = "donor-segment size (cM)", y = "density", colour = NULL, linetype = NULL,
       title = "Simulated cM segment lengths vs the derived gamma law",
       subtitle = sprintf("n=%d segments from the n=1500 BC2S3 sim | derived Gamma KS D=%.3f",
                          nrow(sim_cm), ks_gamma)) +
  theme(legend.position = c(0.72, 0.8), legend.background = element_rect(fill = "white", colour = NA))
```

# Part 3 — cM-space validation of the four callers

Convert each caller's **called Mb segments back to cM** by evaluating the consensus cumulative
map at the segment endpoints and differencing,
$\Delta\text{cM}=M(\text{end\_bp})-M(\text{start\_bp})$ — a *forward* map lookup (per-chromosome
monotone linear interpolation), no inversion. Then compare each caller's cM distribution to the
**map-free reference**: the derived $\text{Gamma}(k,\lambda)$ and, equivalently, the
simulation's own cM lengths.

```{r load-calls}
skim    <- read_csv(params$skim_calls, show_col_types = FALSE)
brb     <- read_csv(params$brb_taxa,   show_col_types = FALSE)
skimbin <- read_csv(params$skimbin,    show_col_types = FALSE)
nilhmm  <- read_csv(params$nilhmm,     show_col_types = FALSE)   # count-based caller, same skim reads
corr    <- read_csv(params$corr, show_col_types = FALSE) %>% filter(type == "NIL")

# scope = full set of NILs called on BOTH platforms (mirrors the Mb intro-size panel)
brb_in  <- corr %>% separate_rows(brbseq_prefix, sep = ";") %>%
  transmute(name = brbseq_prefix, nil = pedigree) %>% filter(name %in% brb$name)
skim_in <- corr %>% separate_rows(skim_prefix, sep = ";") %>%
  transmute(name = skim_prefix, nil = pedigree) %>% filter(name %in% skim$name)
both_nils <- intersect(brb_in$nil, skim_in$nil)

# consensus map -> per-chr monotone Mb->cM interpolator (dedup anchors that share a bp)
mp <- readRDS(params$map) %>% group_by(chr, bp) %>% summarise(cm = mean(cm), .groups = "drop")
mb2cm <- function(chr, bp) {
  out <- numeric(length(bp))
  for (cc in unique(chr)) {
    i <- chr == cc; a <- mp %>% filter(chr == cc) %>% arrange(bp)
    out[i] <- approx(a$bp, a$cm, xout = bp[i], rule = 2)$y   # rule=2 clamps endpoints past the last anchor
  }
  out
}

# donor block = maximal run of donor state (HET or ALT) within (sample, chromosome); size in cM
donor_blocks_cm <- function(seg) seg %>% arrange(name, chr, start_bp) %>% group_by(name, chr) %>%
  mutate(run = cumsum(c(TRUE, diff(state > 0) != 0)), donor = state > 0) %>% filter(donor) %>%
  group_by(name, chr, run) %>%
  summarise(s = min(start_bp), e = max(end_bp), .groups = "drop") %>%
  mutate(cm = mb2cm(chr, e) - mb2cm(chr, s)) %>% filter(cm > 0)

skim_b <- donor_blocks_cm(skim    %>% filter(name %in% skim_in$name[skim_in$nil %in% both_nils])) %>% mutate(src = "Skim-RTI (r=5)")
brb_b  <- donor_blocks_cm(brb     %>% filter(name %in% brb_in$name [brb_in$nil  %in% both_nils])) %>% mutate(src = "BrB-RTI (r=7, ws)")
bin_b  <- donor_blocks_cm(skimbin %>% filter(name %in% skim_in$name[skim_in$nil %in% both_nils])) %>% mutate(src = "Skim-BIN (1 Mb)")
# nilHMM keyed on the skim sample name (same reads as Skim-RTI/Skim-BIN)
nil_b  <- donor_blocks_cm(nilhmm  %>% filter(name %in% skim_in$name[skim_in$nil %in% both_nils])) %>% mutate(src = "nilHMM (counts)")
called <- bind_rows(skim_b, brb_b, bin_b, nil_b)

cat(sprintf("NILs on both platforms: %d | Skim-RTI %d | BrB-RTI %d | Skim-BIN %d | nilHMM %d blocks\n",
            length(both_nils), nrow(skim_b), nrow(brb_b), nrow(bin_b), nrow(nil_b)))
```

```{r ks-table}
ks_vs <- function(x) tibble(
  n          = length(x),
  median_cM  = median(x),
  mean_cM    = mean(x),
  D_vs_Gamma = suppressWarnings(ks.test(x/100, "pgamma", shape = k, rate = lam)$statistic),
  D_vs_sim   = suppressWarnings(ks.test(x, sim_cm$cm)$statistic))

ks_summary <- called %>% group_by(src) %>% group_modify(~ ks_vs(.x$cm)) %>%
  mutate(across(c(median_cM, mean_cM), ~round(.,2)), across(starts_with("D_"), ~round(.,3)))
knitr::kable(ks_summary, caption = "Caller cM segment distributions vs the derived Gamma and the n=1500 sim (KS D).")
```

```{r plot-ecdf, fig.width=8, fig.height=5}
xs2 <- seq(0, quantile(c(called$cm, sim_cm$cm), 0.99), length.out = 400)
ref <- tibble(cm = xs2, F = pgamma(xs2/100, shape = k, rate = lam))

ggplot(called, aes(cm, colour = src)) +
  stat_ecdf(linewidth = 0.9) +
  geom_line(data = ref, aes(cm, F), colour = "black", linewidth = 1.0, linetype = "dashed",
            inherit.aes = FALSE) +
  scale_colour_manual(values = c("Skim-RTI (r=5)" = "purple4", "BrB-RTI (r=7, ws)" = "orange3",
                                 "Skim-BIN (1 Mb)" = "seagreen4", "nilHMM (counts)" = "steelblue")) +
  coord_cartesian(xlim = c(0, quantile(called$cm, 0.99))) +
  labs(x = "donor-segment size (cM)", y = "ECDF", colour = NULL,
       title = "Caller cM segment distributions vs the derived analytic law",
       subtitle = sprintf("dashed black = derived Gamma(k=%.2f, λ=%.1f) | %d NILs on both platforms",
                          k, lam, length(both_nils))) +
  theme(legend.position = c(0.78, 0.3), legend.background = element_rect(fill = "white", colour = NA))
```

```{r plot-density, fig.width=8, fig.height=5}
# On a log10 x-axis the histogram density is per-unit-log10(cM), so both references must be in
# that same space: the sim freqpoly is a ggplot stat (already in-space); the analytic gamma is a
# per-cM density and needs the change-of-variables Jacobian x*ln(10) to match the bars.
ggplot(called, aes(cm, fill = src)) +
  geom_histogram(aes(y = after_stat(density)), position = "identity", alpha = 0.45, bins = 30) +
  geom_freqpoly(data = sim_cm, aes(cm, after_stat(density)), bins = 30,
                colour = "grey25", linewidth = 0.8, inherit.aes = FALSE) +
  geom_line(data = curves %>% filter(grepl("derived", curve)),
            aes(cm, d * cm * log(10)), colour = "black", linewidth = 0.9, inherit.aes = FALSE) +
  scale_fill_manual(values = c("Skim-RTI (r=5)" = "purple4", "BrB-RTI (r=7, ws)" = "orange3",
                               "Skim-BIN (1 Mb)" = "seagreen4", "nilHMM (counts)" = "steelblue")) +
  scale_x_log10() +
  labs(x = "donor-segment size (cM, log scale)", y = "density (per unit log10 cM)", fill = NULL,
       title = "Caller cM segment-size distributions vs the derived law",
       subtitle = "black = derived Gamma (Jacobian-corrected for the log axis) | grey = n=1500 sim cM lengths") +
  theme(legend.position = c(0.01, 0.99), legend.justification = c("left", "top"),
        legend.background = element_rect(fill = "white", colour = NA))
```

## Per-method gamma fits

Each caller's own cM fragment-size distribution, with a **gamma fit by MLE to that caller's
data** (its own $k,\lambda$ in the legend) — not the simulation-derived law. This reads off the
*effective* shape and rate each method's calls imply, and how well a gamma describes each.

```{r per-method-fit, fig.width=7.5, fig.height=8}
fit_gamma <- function(x) {
  f <- suppressWarnings(MASS::fitdistr(x / 100, "gamma")$estimate)   # fit in Morgans
  tibble(k = unname(f["shape"]), lam = unname(f["rate"]))
}
fits <- called %>% group_by(src) %>% group_modify(~ fit_gamma(.x$cm)) %>% ungroup() %>%
  left_join(called %>% group_by(src) %>%
              summarise(xmax = quantile(cm, 0.99), mean_cm = mean(cm), .groups = "drop"),
            by = "src") %>%
  mutate(lab = sprintf("%s — Gamma(k=%.2f, λ=%.1f/M, mean %.0f cM)", src, k, lam, mean_cm))

mle_curves <- fits %>% group_by(src, lab) %>%
  reframe(cm = seq(0, xmax, length.out = 300), d = dgamma(cm / 100, k, lam) / 100)

src_col <- c("Skim-RTI (r=5)" = "purple4", "BrB-RTI (r=7, ws)" = "orange3",
             "Skim-BIN (1 Mb)" = "seagreen4", "nilHMM (counts)" = "steelblue")
pal_lab <- setNames(src_col[fits$src], fits$lab)

ggplot(called, aes(cm)) +
  geom_histogram(aes(y = after_stat(density)), bins = 30,
                 fill = "grey82", colour = "grey55", linewidth = 0.2) +
  geom_line(data = mle_curves, aes(cm, d, colour = lab), linewidth = 1) +
  facet_wrap(~ src, ncol = 1, scales = "free") +
  scale_colour_manual(values = pal_lab, name = "MLE gamma fit (per method)") +
  labs(x = "donor-segment size (cM)", y = "density",
       title = "Per-method cM fragment-size distributions with their own MLE gamma fit") +
  theme(legend.position = "bottom") +
  guides(colour = guide_legend(ncol = 1))
```

Each method's own MLE gamma vs (i) **its own data** — does a gamma describe its shape at all
— and (ii) the **sim-derived benchmark gamma** — how far its calls sit from the neutral model.
The simulation row is the reference (its own gamma *is* the benchmark, so both KS columns equal
its self-fit floor).

```{r per-method-ks-table}
bench_k <- k; bench_lam <- lam   # sim-derived benchmark gamma (from Part 2)
fit_row <- function(x, label) {
  f <- suppressWarnings(MASS::fitdistr(x / 100, "gamma")$estimate)
  tibble(method = label, n = length(x),
         k = unname(f["shape"]), lambda = unname(f["rate"]), mean_cM = mean(x),
         D_own_gamma = suppressWarnings(ks.test(x/100, "pgamma", shape = f["shape"], rate = f["rate"])$statistic),
         D_vs_sim_gamma = suppressWarnings(ks.test(x/100, "pgamma", shape = bench_k, rate = bench_lam)$statistic))
}
ks_methods <- bind_rows(
  fit_row(sim_cm$cm, "Simulation (n=1500)"),
  called %>% group_split(src) %>%
    map_dfr(~ fit_row(.x$cm, .x$src[1]))) %>%
  mutate(across(c(k, lambda, mean_cM), ~round(., 2)),
         across(starts_with("D_"), ~round(., 3)))
knitr::kable(ks_methods,
  caption = "Per-method MLE gamma fit and KS distance, vs the method's own gamma and vs the sim-derived benchmark gamma (k=1.15, λ=4.72/Morgan).")
```

## The Mb-space expectation is lognormal, not gamma

The clean gamma holds **in cM**. Pushed through the empirical, position-dependent genetic map
(ρ(x) varies ~10× along the genome), the **Mb** fragment-size distribution becomes markedly more
skewed (sim Mb CV 1.43, skew 2.08, vs the cM gamma's 0.88, 1.57) and is **best described by a
lognormal, not a gamma** — and there is **no closed-form cM→Mb conversion** (the position-mixture
has no analytic form; this is why the derivation lives in cM). The table compares ways to express
the BC2S3 expectation in Mb.

```{r mb-lognormal}
mb        <- nil_seg$mb[nil_seg$mb > 0]
genome_mb <- sum(read_csv("data/rtiger_50K/seqlengths.csv", show_col_types = FALSE)[[2]]) / 1e6
cm_tot    <- mp %>% group_by(chr) %>% summarise(L = max(cm), .groups = "drop") %>% pull(L) %>% sum()
rbar      <- cm_tot / genome_mb                      # genome-mean rate (cM/Mb)

mom      <- list(shape = mean(mb)^2 / var(mb), rate = mean(mb) / var(mb))
g_mb     <- suppressWarnings(MASS::fitdistr(mb, "gamma", start = mom)$estimate)
ln_mb    <- suppressWarnings(MASS::fitdistr(mb, "lognormal")$estimate)
lam_unif <- lam / 100 * rbar                         # cM gamma rate -> per-Mb under a uniform map

mk <- function(name, params, med, ks)
  tibble(`Mb representation` = name, params = params,
         `median Mb` = round(med, 1), `KS vs sim Mb` = round(ks, 3))
mb_tbl <- bind_rows(
  mk("(A) n=1500 sim Mb segments — the expectation", "empirical", median(mb), 0),
  mk("(A′) lognormal MLE", sprintf("μ=%.2f, σ=%.2f", ln_mb["meanlog"], ln_mb["sdlog"]),
     exp(ln_mb["meanlog"]), suppressWarnings(ks.test(mb, "plnorm", meanlog = ln_mb["meanlog"], sdlog = ln_mb["sdlog"])$statistic)),
  mk("(A″) gamma MLE", sprintf("k=%.2f, rate=%.3f/Mb", g_mb["shape"], g_mb["rate"]),
     qgamma(.5, g_mb["shape"], g_mb["rate"]), suppressWarnings(ks.test(mb, "pgamma", shape = g_mb["shape"], rate = g_mb["rate"])$statistic)),
  mk("(B) uniform-rate scaled cM gamma", sprintf("k=%.2f (same shape), rate=%.3f/Mb", k, lam_unif),
     qgamma(.5, k, lam_unif), suppressWarnings(ks.test(mb, "pgamma", shape = k, rate = lam_unif)$statistic)))
knitr::kable(mb_tbl, caption = sprintf(
  "BC2S3 expectation in Mb (genome %.0f Mb, map %.0f cM, mean rate %.2f cM/Mb). The cM gamma does NOT push forward to a gamma in Mb; lognormal fits best. (B) the naive uniform-rate scaling gets the mean right but the median/tail wrong.",
  genome_mb, cm_tot, rbar))
```

**Convention.** The **n=1500 BC2S3 simulation** (`results/nil_segments.csv`, `mb` column) **is the
BC2S3 expectation in Mb** — use it directly as the reference (Null) in all Mb introgression-size
plots. Do **not** substitute the uniform-rate-scaled gamma (row B: KS ≈ 0.22; right mean, wrong
median/tail). The smooth parametric stand-in, if needed, is the lognormal (row A′), not a gamma.
RTIGER rigidity (`r`) calibration targets this same distribution — the BrB-RTI KS sweep already
does (`ks_sweep_brbseq_wideseq.R` keys on the `mb` column); Skim-RTI's `r=5` should be reconciled
against the same reference.

# Interpretation & caveats

- **The derivation is validated:** the BC2S3 recombination model predicts a near-exponential
  gamma ($k\gtrsim1$, $\lambda\lesssim5$/Morgan), and the simulation's own cM lengths match it to
  KS $D\approx`r round(ks_gamma,3)`$ — so the analytic law is a faithful, map-free stand-in for
  the n=1500 reference, now available in closed form.
- **Caller comparison in cM** (table above) mirrors the Mb panel but against the analytic
  reference. Deviations read the same way: a caller that over-fragments donor blocks shifts mass
  to short cM; one that under-resolves (Skim-BIN's 1 Mb bins) loses the short tail.
- **Conserved marker order is load-bearing.** The cM conversion uses the consensus $\rho(x)$;
  for any introgression with local rearrangement the map is wrong and $\Delta$cM is biased. This
  assumption is *heavier* here than in the Mb panel.
- **Pericentromeric pile-up.** Recombination deserts map large Mb spans to ≈0 cM, so a chunk of
  centromeric blocks collapse near cM 0 — map physics, not caller behavior.
- **Keep Mb for tuning `r`.** The caller resolution limit is defined in Mb; cM smears it through
  the map. This panel is the *validation* view; the [Mb panel](skim_brbseq_skimbin.html) remains
  the calibration space. Agreement across both is the robustness check.