---
title: "Ancestry callers compared — GBTS, RTIGER skim, and Gaussian-mixture HMM"
author: "Fausto Rodríguez-Zapata"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    df-print: paged
    embed-resources: true
execute:
  warning: false
  message: false
params:
  mb_calls:   "results/sim_calibration/gbts_calls/SNP_wsfilt/calls_common_schema.csv"  # GBTS real counts, wsfilt, r=2
  mb_counts:  "results/sim_calibration/gbts_calls/SNP_wsfilt/counts"                    # wsfilt marker grid (9,157)
  skim_calls: "data/rtiger_50K/calls_taxa_r5.csv"                                       # skim RTIGER, per-taxon r=5
  gmm_calls:  "data/bzeaseq_ancestry/gmm_hmm_segments_16lines.tsv"                      # BzeaSeq Gaussian-mixture HMM (skim)
  sample_map: "data/molbreeding_45k_sample_map.tsv"
  seqlengths: "data/rtiger_50K/seqlengths.csv"
  seed:       1
---

**Repo:** [sawers-rellan-labs/zealtiger](https://github.com/sawers-rellan-labs/zealtiger)

Three ancestry callers on the **same 16 BZea lines**, painted side by side:

- **GBTS r=2** — MolBreeding ~110× target-capture, RTIGER at the truth-derived
  rigidity (`docs/15-rigidity-and-scoring.md`).
- **RTIGER skim r=5** — SNP50K ~0.4× skim, RTIGER at the truth-derived rigidity.
- **GMM-HMM skim** — the BzeaSeq caller on the **same skim data** (1 Mb bins,
  Gaussian mixture on `ALT_FREQ`, then HMM-smoothed; `BzeaSeq/docs/call_ancestry.Rmd`).

This is the companion to the two-way `molbreeding_vs_skim_calibrated.qmd` (which is
GBTS vs RTIGER skim only). The point here is to add an **independent skim caller**
and ask whether the GBTS-vs-skim gap is a RTIGER artifact or a platform effect.
Palette follows the BzeaSeq convention: `REF` gold, `HET` springgreen4, `ALT`
purple4. Calls are segments on B73 NAM v5 (0/REF, 1/HET, 2/ALT). The two B73-bulk
checks (`PN4_SID326`, `PN6_SID576`) have no RTIGER-skim lane (GBTS-only).

## Setup

```{r setup}
library(tidyverse)
set.seed(params$seed)
pal <- c(REF = "gold", HET = "springgreen4", ALT = "purple4")

mb   <- read_csv(params$mb_calls,   show_col_types = FALSE)   # GBTS wsfilt, r=2
skim <- read_csv(params$skim_calls, show_col_types = FALSE)   # skim RTIGER, r=5
smap <- read_tsv(params$sample_map, show_col_types = FALSE)
gmm  <- read_tsv(params$gmm_calls,  show_col_types = FALSE) %>%   # GMM-HMM skim
  transmute(name = sample, chr = as.integer(sub("chr","",chrom)),
            start_bp = start, end_bp = end, state = state_chr <- genotype)

checks <- c("PN4_SID326", "PN6_SID576")

# per-sample donor dosage (bp-weighted state/2), used to order the lines
dosage <- function(seg) seg %>% mutate(len = end_bp - start_bp + 1) %>%
  group_by(name) %>% summarise(d = sum(len * state/2) / sum(len), .groups = "drop")
gbts_d <- dosage(mb)
ord <- c(checks, gbts_d %>% filter(!name %in% checks) %>% arrange(d) %>% pull(name))

# LowQuality samples (sample_map QC flag) get a bold facet-strip label (markdown)
lowq <- smap$sample_id[smap$quality == "LowQuality" & !is.na(smap$quality)]
name_lab <- function(x) if_else(x %in% lowq, paste0("**", x, "**"), x)

# normalise each call set to (name, chr, start_bp, end_bp, state=REF/HET/ALT, method)
as_paint <- function(seg, method, numeric_state = TRUE) {
  s <- if (numeric_state) factor(seg$state, 0:2, c("REF","HET","ALT")) else factor(seg$state, c("REF","HET","ALT"))
  tibble(name = seg$name, chr = seg$chr, start_bp = seg$start_bp, end_bp = seg$end_bp,
         state = s, method = method)
}
```

## Chromosome painting — all three callers

Three lanes per sample: **top = GBTS r=2, middle = RTIGER skim r=5, bottom =
GMM-HMM skim**. Lines ordered by GBTS donor dosage (B73 checks pinned at the top);
`LowQuality` lines in **bold**.

```{r painting-3way, fig.width=12, fig.height=20}
paint3 <- bind_rows(
  as_paint(mb,   "GBTS r=2"),
  as_paint(skim %>% filter(name %in% unique(mb$name)), "RTIGER skim r=5"),
  as_paint(gmm,  "GMM-HMM skim", numeric_state = FALSE)) %>%
  mutate(chr    = factor(chr, levels = 1:10, labels = paste0("chr", 1:10)),
         name   = factor(name_lab(name), levels = name_lab(ord)),
         method = factor(method, c("GBTS r=2","RTIGER skim r=5","GMM-HMM skim")),
         ylo = case_when(method == "GBTS r=2" ~ 0.69, method == "RTIGER skim r=5" ~ 0.35, TRUE ~ 0.00),
         yhi = case_when(method == "GBTS r=2" ~ 1.00, method == "RTIGER skim r=5" ~ 0.66, TRUE ~ 0.32))

ggplot(paint3, aes(xmin = start_bp/1e6, xmax = end_bp/1e6,
                   ymin = ylo, ymax = yhi, fill = state)) +
  geom_rect() +
  facet_grid(name ~ chr, scales = "free_x", space = "free_x", switch = "y") +
  scale_fill_manual(values = pal, drop = FALSE) +
  scale_x_continuous(labels = scales::unit_format(unit = "", scale = 1)) +
  labs(title = "Ancestry calls — 16 lines, 3 callers per sample",
       subtitle = "per sample top→bottom: GBTS r=2 / RTIGER skim r=5 / GMM-HMM skim | gold REF, green HET, purple ALT | bold = LowQuality",
       x = "Position (Mb)", fill = "Genotype") +
  ggpubr::theme_classic2(base_size = 9) +
  theme(legend.position = "bottom",
        strip.background = element_blank(),
        strip.text.x = element_text(face = "bold"),
        strip.text.y.left = ggtext::element_markdown(angle = 0, hjust = 1, size = 7),
        panel.spacing.x = unit(0.05, "lines"),
        panel.spacing.y = unit(0.05, "lines"),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        axis.title.y = element_blank(), axis.line.y = element_blank(),
        axis.text.x = element_text(size = 6))
```

**Read:** the two skim lanes (RTIGER, GMM-HMM) are near-copies of each other and
both sit **below** the GBTS lane. The skim/GBTS gap is a platform effect
reproduced by two independent skim callers, not a single-caller artifact.
`PN4_SID330` is the lone sample where the two skim lanes agree with each other but
**disagree in pattern** with GBTS — the sample-identity swap flagged in
`molbreeding_vs_skim_calibrated.qmd`.

## Donor genome-fraction — the three callers quantified

```{r donor-fraction}
frac <- function(seg, lab, numeric_state = TRUE) {
  st <- if (numeric_state) seg$state > 0 else seg$state != "REF"
  het <- if (numeric_state) seg$state == 1 else seg$state == "HET"
  seg %>% mutate(mb = (end_bp - start_bp)/1e6, donor = st, is_het = het) %>%
    group_by(name) %>%
    summarise(!!paste0(lab,"_donor") := sum(mb[donor])/sum(mb),
              !!paste0(lab,"_het")   := sum(mb[is_het])/sum(mb), .groups = "drop")
}
tab <- frac(mb, "gbts") %>%
  left_join(frac(skim %>% filter(name %in% unique(mb$name)), "rtiger"), by = "name") %>%
  left_join(frac(gmm, "gmm", numeric_state = FALSE), by = "name") %>%
  arrange(gbts_donor)
knitr::kable(tab, digits = 3)

cat("\nmeans (NILs only, B73 checks dropped):\n")
tab %>% filter(!name %in% checks) %>%
  summarise(across(where(is.numeric), ~round(mean(., na.rm = TRUE), 3))) %>%
  knitr::kable()
```

**Read:** RTIGER-skim and GMM-HMM-skim donor fractions are near-identical per
sample (means donor 0.093 vs 0.081, het 0.062 vs 0.059) and both sit below GBTS
(donor 0.141, het 0.116). Two independent skim callers reproduce the same
under-call relative to GBTS — so the gap is platform/coverage (and/or GBTS r=2
over-calling), not a RTIGER quirk. `PN4_SID330`: both skim callers ~0.08 donor vs
GBTS 0.29 — independent corroboration that this line's GBTS and skim DNA are not
the same sample.

## Two-way — GBTS vs GMM-HMM skim

The GBTS-vs-skim comparison through the GMM-HMM caller alone (top = GBTS r=2,
bottom = GMM-HMM skim), the GMM-HMM analog of the RTIGER two-way figure in
`molbreeding_vs_skim_calibrated.qmd`.

```{r painting-2way-gmm, fig.width=12, fig.height=16}
paint2 <- bind_rows(
  as_paint(mb, "GBTS r=2"),
  as_paint(gmm, "GMM-HMM skim", numeric_state = FALSE)) %>%
  mutate(chr    = factor(chr, levels = 1:10, labels = paste0("chr", 1:10)),
         name   = factor(name_lab(name), levels = name_lab(ord)),
         ylo = if_else(method == "GBTS r=2", 0.52, 0.00),
         yhi = if_else(method == "GBTS r=2", 1.00, 0.48))

ggplot(paint2, aes(xmin = start_bp/1e6, xmax = end_bp/1e6,
                   ymin = ylo, ymax = yhi, fill = state)) +
  geom_rect() +
  facet_grid(name ~ chr, scales = "free_x", space = "free_x", switch = "y") +
  scale_fill_manual(values = pal, drop = FALSE) +
  scale_x_continuous(labels = scales::unit_format(unit = "", scale = 1)) +
  labs(title = "Ancestry calls — 16 lines, GBTS r=2 (top) vs GMM-HMM skim (bottom)",
       subtitle = "gold REF, green HET, purple ALT | bold = LowQuality",
       x = "Position (Mb)", fill = "Genotype") +
  ggpubr::theme_classic2(base_size = 9) +
  theme(legend.position = "bottom",
        strip.background = element_blank(),
        strip.text.x = element_text(face = "bold"),
        strip.text.y.left = ggtext::element_markdown(angle = 0, hjust = 1, size = 7),
        panel.spacing.x = unit(0.05, "lines"),
        panel.spacing.y = unit(0.05, "lines"),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        axis.title.y = element_blank(), axis.line.y = element_blank(),
        axis.text.x = element_text(size = 6))
```

## Notes

- GMM-HMM calls staged from the BzeaSeq pipeline output
  `all_samples_ancestry_segments.bed` (the lab-meeting call set) to
  `data/bzeaseq_ancestry/gmm_hmm_segments_16lines.tsv`; not recomputed here.
- The GMM-HMM and RTIGER-skim callers run on the **same skim data**; their close
  agreement is an independent check that the GBTS-vs-skim divergence is not a
  caller artifact.