---
title: "MolBreeding GBTS vs SNP50K skim — calibrated ancestry-call comparison"
author: "Fausto Rodríguez-Zapata"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    df-print: paged
    embed-resources: true
execute:
  warning: false
  message: false
params:
  mb_calls:   "results/sim_calibration/gbts_calls/SNP_wsfilt/calls_common_schema.csv"  # GBTS real counts, wsfilt, r=2
  mb_counts:  "results/sim_calibration/gbts_calls/SNP_wsfilt/counts"                    # wsfilt marker grid (9,157)
  skim_calls: "data/rtiger_50K/calls_taxa_r5.csv"                                       # skim, per-taxon r=5
  sample_map: "data/molbreeding_45k_sample_map.tsv"
  nil_seg:    "results/nil_segments.csv"     # n=1500 simulation introgression segments (NULL)
  nil_co:     "results/nil_crossovers.csv"   # n=1500 simulation crossover counts (NULL)
  seqlengths: "data/rtiger_50K/seqlengths.csv"
  seed:       1
---

**Repo:** [sawers-rellan-labs/zealtiger](https://github.com/sawers-rellan-labs/zealtiger)

Calibrated cross-validation of ancestry source **#2 (SNP50K skim, ~0.4×)** against
source **#5 (MolBreeding GBTS, ~110×)** for the BZea NILs genotyped on both
platforms. This is the **calibrated** successor to the frozen direct comparison
(`docs/molbreeding_vs_skim.html`).

**What changed vs the frozen comparison.** The frozen notebook compared both
platforms at a single borrowed **r=8**. Here each platform is called at its
**truth-derived** rigidity — **skim r=5, GBTS r=2** (FP+FN argmin vs the simulated
truth; methodology `docs/15-rigidity-and-scoring.md`), because the two coverage
regimes have different physical resolution (`docs/14`). The **null** is the
**n=1500 BC2S3 simulation** (consensus map, no selection) — a modeled expectation,
**not ground truth**: divergence from it is an observation. The per-platform
calibration is done in `skim_sim_calibration.qmd` and `gbts_sim_calibration.qmd`.

All sites are on the **wideseq-filtered** high-confidence panel (9,157 v5 SNPs,
consistent REF=B73 / ALT=teosinte polarity). All 16 GBTS samples are kept (14
teosinte NILs + 2 B73-bulk checks); the 14 NILs are shared with the skim. Calls
are RTIGER segments on B73 NAM v5 (`name, chr, start_bp, end_bp, state`;
0 = recurrent/B73, 1 = het, 2 = homozygous donor). Sample IDs are joined to GBTS
wells via `data/molbreeding_45k_sample_map.tsv` (well_id → PN#_SID#); the calls
already carry the PN#_SID# names.

## Setup

```{r setup}
library(tidyverse); library(GenomicRanges); library(S4Vectors)
source("R/06_score_rtiger.R")
source("R/07_single_locus.R")
set.seed(params$seed)

mb      <- read_csv(params$mb_calls,   show_col_types = FALSE)   # GBTS wsfilt, r=2
skim    <- read_csv(params$skim_calls, show_col_types = FALSE)   # skim, per-taxon r=5
smap    <- read_tsv(params$sample_map, show_col_types = FALSE)
sl      <- read_csv(params$seqlengths, show_col_types = FALSE)
nil_seg <- read_csv(params$nil_seg,    show_col_types = FALSE)   # NULL: introgression segments
nil_co  <- read_csv(params$nil_co,     show_col_types = FALSE)   # NULL: crossover counts

p0 <- single_locus_p0(n_bc = 2L, n_self = 3L)  # neutral breeding-scheme expectation

# common wideseq marker grid, from the filtered RTIGER count files
cf <- read_tsv(list.files(params$mb_counts, full.names = TRUE)[1],
               col_names = c("chr_label","bp","rb","rc","ab","ac"), show_col_types = FALSE)
grid    <- tibble(chr = as.integer(sub("chr","",cf$chr_label)), bp = cf$bp)
spacing <- sum(sl$len) / nrow(grid)

# samples present on both platforms (the 2 B73 checks have no skim NIL match)
samples <- intersect(unique(mb$name), unique(skim$name))
cat(sprintf("MolBreeding GBTS samples: %d (all kept) | shared with skim: %d\n",
            n_distinct(mb$name), length(samples)))
cat(sprintf("wideseq markers: %d (~%.0f kb spacing)\n", nrow(grid), spacing/1e3))
cat(sprintf("resolution floor — skim r=5 ≈ %.1f Mb | GBTS r=2 ≈ %.1f Mb\n",
            5*spacing/1e6, 2*spacing/1e6))

# donor blocks from a STATE-segment table (skim, GBTS): merge runs of state>0
donor_blocks <- function(seg) seg %>% arrange(name, chr, start_bp) %>% group_by(name, chr) %>%
  mutate(run = cumsum(c(TRUE, diff(state > 0) != 0)), donor = state > 0) %>% filter(donor) %>%
  group_by(name, chr, run) %>% summarise(size_mb = (max(end_bp)-min(start_bp))/1e6, .groups = "drop")
co_count <- function(seg) seg %>% arrange(name, chr, start_bp) %>% group_by(name, chr) %>%
  summarise(tr = sum(abs(diff(as.integer(state > 0)))), .groups = "drop") %>%
  group_by(name) %>% summarise(COs = sum(tr), .groups = "drop")
state_at <- function(df, s) {
  d <- df[df$name == s, ]
  d$state[findOverlaps(GRanges(grid$chr, IRanges(grid$bp, width = 1)),
                       GRanges(d$chr, IRanges(d$start_bp, d$end_bp)), select = "first")]
}
```

## Donor dosage & the B73-bulk control

Per-sample donor dosage (bp-weighted `state/2`). The neutral BC2S3 expectation
≈ 0.125; a flipped REF/ALT polarity would read ≈ 0.875. The two B73-bulk checks
(`PN4_SID326`, `PN6_SID576`) should read ≈ 0 — a built-in validity test that needs
no external truth.

```{r dosage}
dosage <- function(seg) seg %>% mutate(len = end_bp - start_bp + 1) %>%
  group_by(name) %>% summarise(dosage = sum(len * state/2) / sum(len), .groups = "drop")
checks <- c("PN4_SID326", "PN6_SID576")
dose <- dosage(mb) %>% dplyr::rename(gbts_r2 = dosage) %>%
  left_join(dosage(skim) %>% dplyr::rename(skim_r5 = dosage), by = "name") %>%
  mutate(is_B73_check = name %in% checks) %>% arrange(gbts_r2)
dose
cat(sprintf("mean donor dosage — GBTS wsfilt r=2 %.3f | skim r=5 (shared) %.3f | neutral p0 %.3f\n",
            mean(dose$gbts_r2[!dose$is_B73_check]),
            mean(dose$skim_r5[!is.na(dose$skim_r5)]), p0["ALT"] + 0.5*p0["Het"]))
cat(sprintf("B73-bulk checks dosage (GBTS): %s — expect ≈ 0\n",
            paste(sprintf("%s=%.3f", checks,
                          dose$gbts_r2[match(checks, dose$name)]), collapse = ", ")))
```

**Read:** polarity is correct (dosage far below the ~0.875 of a flipped panel) and
the B73-bulk checks read ≈ 0 — the built-in control passes. Both platforms read
donor dosage **well below** the neutral 0.125, the donor-homozygote deficit the
calibrated three-`f_i` analysis attributes to biology (`skim_sim_calibration.qmd`,
`gbts_sim_calibration.qmd`).

## State on the common grid

```{r state-grid}
paired <- map_dfr(samples, function(s) {
  a <- state_at(mb, s); k <- state_at(skim, s); ok <- !is.na(a) & !is.na(k)
  tibble(name = s, gbts = a[ok], skim = k[ok])
})
```

## Confusion matrix — GBTS vs skim

Rows = GBTS state (wideseq-filtered, r=2), cols = skim (per-taxon r=5); cell =
count (row %). Common wsfilt grid (`r nrow(grid)` sites).

```{r confusion}
cm <- table(gbts = factor(paired$gbts, 0:2), skim = factor(paired$skim, 0:2))
overall <- sum(diag(cm)) / sum(cm)
prop <- sweep(cm, 1, rowSums(cm), "/")
disp <- matrix(sprintf("%d (%.1f%%)", cm, 100*prop), nrow = 3,
               dimnames = list(`gbts \\ skim` = c("0 recurrent","1 het","2 donor"),
                               c("0 recurrent","1 het","2 donor")))
knitr::kable(disp, align = "r",
             caption = sprintf("GBTS r=2 (rows) vs skim r=5 (cols): count (row %%). Overall concordance %.3f.", overall))
dp <- paired$gbts %in% c(1, 2)
cat(sprintf("GBTS donor-present markers: %d | skim also donor-present: %.1f%%\n",
            sum(dp), 100*mean(paired$skim[dp] %in% c(1,2))))
cat(sprintf("GBTS homozygous-donor markers: %d | skim also hom-donor: %.1f%%\n",
            sum(paired$gbts == 2), 100*mean(paired$skim[paired$gbts == 2] == 2)))
```

**Read:** marker-level concordance is high; the disagreements concentrate where
GBTS calls **het** and the skim calls **recurrent** — the ~0.4× skim's het
under-call, the direction the calibrated benchmarks predict.

## Segment scale & crossover count — GBTS vs skim vs null

Null = the **n=1500 simulation** introgression segments (`nil_segments.csv`) and
crossover counts (`nil_crossovers.csv`). For the null, "donor blocks" are the
simulated introgression segments directly (no REF segments in that table).

```{r seg-scale}
metrics_calls <- function(seg, lab, nsamp) {
  db <- donor_blocks(seg)
  tibble(source = lab,
         donor_blocks_per_sample = round(nrow(db)/nsamp, 1),
         median_donor_block_Mb   = round(median(db$size_mb), 2),
         median_COs              = median(co_count(seg)$COs))
}
mb_s   <- mb   %>% filter(name %in% samples)
skim_s <- skim %>% filter(name %in% samples)
null_row <- tibble(source = "simulation (null, n=1500)",
                   donor_blocks_per_sample = round(nrow(nil_seg)/n_distinct(nil_seg$nil_id), 1),
                   median_donor_block_Mb   = round(median(nil_seg$mb), 2),
                   median_COs              = median(nil_co$n_co))
bind_rows(
  metrics_calls(mb_s,   "GBTS wsfilt (r=2)", length(samples)),
  metrics_calls(skim_s, "skim (r=5)",        length(samples)),
  null_row) %>%
  knitr::kable()
```

**Read:** GBTS resolves more, smaller donor blocks and more crossovers than the
coverage-limited skim — the high-coverage platform recovers fine structure the
skim merges. Both call **fewer** crossovers than the null simulation expects (the
caller's segment-merging floor vs the modeled recombination rate); the null is an
expectation, not truth.

## Introgression-size distribution

```{r intro-sizes}
gbts_b <- donor_blocks(mb_s)   %>% mutate(src = "GBTS wsfilt (r=2)")
skim_b <- donor_blocks(skim_s) %>% mutate(src = "skim (r=5)")
null_b <- nil_seg %>% transmute(size_mb = mb)
ggplot(bind_rows(gbts_b, skim_b), aes(x = size_mb, fill = src)) +
  geom_histogram(aes(y = after_stat(density)), position = "identity", alpha = 0.5, bins = 30) +
  geom_freqpoly(data = null_b, aes(x = size_mb, y = after_stat(density)),
                bins = 30, colour = "grey20", linewidth = 0.9, inherit.aes = FALSE) +
  scale_fill_manual(values = c("GBTS wsfilt (r=2)" = "forestgreen", "skim (r=5)" = "purple4")) +
  scale_x_log10() +
  labs(x = "introgression size (Mb, log scale)", y = "density", fill = NULL,
       title = "Introgression-size distribution (calibrated r; 14 shared NILs)",
       subtitle = "grey line = n=1500 simulation (null expectation)") + theme_bw()
```

## Donor-call composition — het vs homozygous donor

Per-sample het and homozygous-donor **fraction** of the common grid, by platform.
The top reference row is the **neutral breeding-scheme p₀**
(Het `r sprintf("%.5f", p0["Het"])`, hom-donor `r sprintf("%.5f", p0["ALT"])`;
`single_locus_p0(2,3)`) — the expectation under no selection.

```{r donor-composition, fig.height=5}
comp <- bind_rows(
  paired %>% transmute(name, platform = "GBTS wsfilt (r=2)", state = gbts),
  paired %>% transmute(name, platform = "skim (r=5)",        state = skim)) %>%
  filter(state %in% c(1, 2)) %>% count(name, platform, state) %>%
  left_join(count(paired, name, name = "tot"), by = "name") %>%
  mutate(pct = n/tot, state = factor(state, c(1,2), c("het","homozygous donor")))

# p0 reference row, drawn at the top of both facets
p0_ref <- expand_grid(platform = c("GBTS wsfilt (r=2)", "skim (r=5)"),
                      tibble(state = factor(c("het","homozygous donor"),
                                            c("het","homozygous donor")),
                             pct = c(p0["Het"], p0["ALT"]))) %>%
  mutate(name = "p0 (breeding scheme)")

ord <- paired %>% group_by(name) %>% summarise(d = mean(gbts)/2) %>% arrange(d) %>% pull(name)
comp_all <- bind_rows(comp, p0_ref) %>%
  mutate(name = factor(name, levels = c(ord, "p0 (breeding scheme)")))
ggplot(comp_all, aes(y = name, x = pct, fill = state)) +
  geom_col(position = "stack") + facet_wrap(~ platform) +
  scale_fill_manual(values = c("het" = "forestgreen", "homozygous donor" = "purple4")) +
  scale_x_continuous(labels = scales::percent) +
  labs(x = "% of genome called donor", y = NULL, fill = NULL,
       title = "Donor calls as % of genome (recurrent not drawn)",
       subtitle = "top row = neutral breeding-scheme p0 (no selection)") + theme_bw()
```

**Read:** nearly every NIL sits **above** the neutral p₀ on het and **below** it on
homozygous donor — the heterozygote-excess / donor-homozygote-deficit signal,
visible on both platforms and **stronger on GBTS** (which resolves het the skim
under-calls). The calibrated three-`f_i` decomposition
(`skim_sim_calibration.qmd`, `gbts_sim_calibration.qmd`) attributes this to real
biology, not method bias.

## Findings

- **Calibrated rigidity, per platform.** Skim r=5, GBTS r=2 (truth-derived;
  `docs/15`), not the frozen shared r=8. The two coverage regimes have different
  physical resolution, so a shared r over-fragments GBTS (`docs/14`).
- **Polarity & controls check out.** Donor dosage is low (BC2S3-like, far from a
  flipped 0.875) and the two B73-bulk checks read ≈ 0 — a built-in validity test
  passed without external truth.
- **GBTS resolves more het and finer structure** than the ~0.4× skim, the expected
  direction; the disagreements concentrate at GBTS-het / skim-recurrent sites.
- **Both platforms sit off the neutral p₀** — het excess, donor-homozygote deficit,
  donor dosage ~0.07 vs 0.125. The simulation (n=1500) is the **null** expectation,
  not truth; the calibrated benchmarks license attributing the deviation to biology
  (`docs/12-het-excess-diagnosis.md`).
```