---
title: "MolB-RTI vs Skim-RTI vs BrB-RTI — ancestry-call concordance (real counts, shared r=8)"
author: "Fausto Rodríguez-Zapata"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    df-print: paged
    embed-resources: true
execute:
  warning: false
  message: false
params:
  mb_calls:   "data/molbreeding_targetseq/SNP_wsfilt/calls_common_schema.csv"  # target-seq real counts, wideseq-filtered, r=8, all 16
  mb_full:    "data/molbreeding_targetseq/SNP/calls_common_schema.csv"         # target-seq real counts, all 20,347 v5 SNPs, r=8
  skim_calls: "data/rtiger_50K/calls_taxa_r8.csv"                         # skim, per-taxon r=8
  brb_calls:  "data/brbseq/calls_common_schema.csv"                       # BRB-seq 3' RNA-seq real counts, r=8 (no wsfilt)
  sample_map: "data/molbreeding_45k_sample_map.tsv"
  mb_counts:  "data/molbreeding_targetseq/SNP_wsfilt/counts"                   # filtered marker grid (RTIGER 6-col)
  sim_seg:    "results/rtiger_45k_benchmark/truth_segments.csv"           # consensus-map simulation (NULL)
  seqlengths: "data/rtiger_50K/seqlengths.csv"
---

**Repo:** [sawers-rellan-labs/zealtiger](https://github.com/sawers-rellan-labs/zealtiger)

Cross-validation of ancestry source **#2 (SNP50K skim, ~0.4×)** against source **#5
(MolBreeding target sequencing)** for the BZea NILs genotyped on both platforms, with a third
real-count platform — **BRB-seq 3′ bulk RNA-seq (~1.5–3×)** — layered in as an
additional cross-check on the NILs it shares.

**BRB-seq layer (source #6).** GATK `CollectAllelicCounts` over the brbseq
joint-called SNP sites gives per-sample `REF_COUNT`/`ALT_COUNT` with REF = B73 NAM
v5 and ALT = observed alternate, so — like the skim and target sequencing — it is fed to RTIGER
as **real allele counts** (`fit_rtiger_brbseq.R`, r = 8, MLE, no wsfilt: the brbseq
sites are used as-is, since RNA-seq covers only expressed regions and overlaps the
wideseq donor panel at too few markers to filter). BRB-seq run_full sequenced plates
PN1–PN4 (SID 1–384), so it overlaps the comparison only on the **8 PN4 samples**
(7 teosinte NILs + the `PN4_SID326` B73-bulk check); the PN6/PN12 NILs have no
BRB-seq counterpart here. BRB-seq sections below are therefore restricted to those
8 samples, while the target-sequencing-vs-skim comparison keeps all 16.

**This is the real-count pipeline.** MolBreeding source #5 is **target-capture
sequencing** (~110× depth, ~0.2% missing), *not* a hybridization array — so it is fed to
RTIGER as **real allele counts** (`REF_COUNT`/`ALT_COUNT` from the GATK table), exactly
like the skim, rather than the 0/1/2 pseudo-counts of the earlier array-style attempt.

**Two design choices:**

1. **No line-QC filter.** The array-era "Crappy"/`Good` split was driven by array
   het-rate; with real ~110× counts every sample is clean (missing ~0.2%), so **all 16
   samples are kept** — 14 teosinte NILs + 2 B73-bulk checks. The checks are a built-in
   control (they should read ≈ pure recurrent).
2. **SNP filter (`wsfilt`).** Keep only markers overlapping the **wideseq
   teosinte-vs-B73 panel** (9,157 of the 20,347 lifted v5 SNPs are donor-informative,
   with consistent REF=B73 / ALT=teosinte polarity across sources). The full 20,347-SNP
   fit is shown alongside for reference.

**Rigidity r = 8 for both platforms** — matched to the skim's per-taxon r=8 (Source 2),
so the comparison uses one consistent rigidity rather than ad-hoc per-platform tuning.
Calls are RTIGER segments on B73 NAM v5 (`name, chr, start_bp, end_bp, state`;
0 = recurrent/B73, 1 = het, 2 = homozygous donor).

> The consensus-map BC2S3 **simulation** (source #1) is a **null expectation** under the
> modeled recombination process (MaizeGDB consensus map, ~1,783 cM; `m=10` interference;
> BC2S3 pedigree), **not ground truth**.

**Naming.** Call sets are labelled `Source-Caller`. Sources: **Skim** (SNP50K
~0.4× skim), **MolB** (MolBreeding target-capture sequencing, ~110×), **BrB** (BRB-seq 3′
RNA-seq). Caller here is **RTI** = RTIGER, run on all three at a shared **r=8**. MolB
has two RTIGER marker sets: **MolB-RTI (wsfilt)** (wideseq donor panel, 9,157 SNPs;
the primary) and **MolB-RTI (full)** (all 20,347 lifted v5 SNPs). The consensus-map
simulation is **Null**. (A 1 Mb-bin GMM-HMM caller, **BIN**, is compared elsewhere;
it does not appear here.)

## Setup

```{r setup}
library(tidyverse); library(GenomicRanges); library(S4Vectors)
source("R/06_score_rtiger.R")

mb      <- read_csv(params$mb_calls,   show_col_types = FALSE)   # target-seq wsfilt, r=8
mb_full <- read_csv(params$mb_full,    show_col_types = FALSE)   # target-seq full,  r=8
skim    <- read_csv(params$skim_calls, show_col_types = FALSE)   # skim, per-taxon r=8
brb     <- read_csv(params$brb_calls,  show_col_types = FALSE)   # BRB-seq RNA, r=8 (no wsfilt)
smap    <- read_tsv(params$sample_map, show_col_types = FALSE)
sl      <- read_csv(params$seqlengths, show_col_types = FALSE)
sim_seg <- read_csv(params$sim_seg,    show_col_types = FALSE)

# common wideseq marker grid, from the filtered RTIGER count files
cf <- read_tsv(list.files(params$mb_counts, full.names = TRUE)[1],
               col_names = c("chr_label","bp","rb","rc","ab","ac"), show_col_types = FALSE)
grid    <- tibble(chr = as.integer(sub("chr","",cf$chr_label)), bp = cf$bp)
spacing <- sum(sl$len) / nrow(grid)

# samples present on both platforms (the 2 B73 checks have no skim NIL match)
samples <- intersect(unique(mb$name), unique(skim$name))
# BRB-seq overlaps only the PN4 subset (run_full = plates PN1-PN4)
brb_samples <- intersect(unique(brb$name), unique(mb$name))
cat(sprintf("MolB samples: %d (all kept, no QC) | shared with Skim: %d\n",
            n_distinct(mb$name), length(samples)))
cat(sprintf("BrB samples: %d | shared with MolB: %d | also with Skim: %d\n",
            n_distinct(brb$name), length(brb_samples),
            length(intersect(brb_samples, samples))))
cat(sprintf("wideseq markers: %d (~%.0f kb spacing; r=8 detection limit ≈ %.1f Mb)\n",
            nrow(grid), spacing/1e3, 8*spacing/1e6))

donor_blocks <- function(seg) seg %>% arrange(name, chr, start_bp) %>% group_by(name, chr) %>%
  mutate(run = cumsum(c(TRUE, diff(state > 0) != 0)), donor = state > 0) %>% filter(donor) %>%
  group_by(name, chr, run) %>% summarise(size_mb = (max(end_bp)-min(start_bp))/1e6, .groups = "drop")
co_count <- function(seg) seg %>% arrange(name, chr, start_bp) %>% group_by(name, chr) %>%
  summarise(tr = sum(abs(diff(as.integer(state > 0)))), .groups = "drop") %>%
  group_by(name) %>% summarise(COs = sum(tr), .groups = "drop")
state_at <- function(df, s) {
  d <- df[df$name == s, ]
  d$state[findOverlaps(GRanges(grid$chr, IRanges(grid$bp, width = 1)),
                       GRanges(d$chr, IRanges(d$start_bp, d$end_bp)), select = "first")]
}
```

## Donor dosage & the B73-bulk control

Per-sample donor dosage (bp-weighted `state/2`). BC2S3 expectation ≈ 0.125; a flipped
REF/ALT polarity would read ≈ 0.875. The two B73-bulk checks (`PN4_SID326`, `PN6_SID576`)
should read ≈ 0 — a built-in validity test that needs no external truth.

```{r dosage}
dosage <- function(seg) seg %>% mutate(len = end_bp - start_bp + 1) %>%
  group_by(name) %>% summarise(dosage = sum(len * state/2) / sum(len), .groups = "drop")
checks <- c("PN4_SID326", "PN6_SID576")
dose <- dosage(mb) %>% dplyr::rename(molb_wsfilt = dosage) %>%
  left_join(dosage(mb_full) %>% dplyr::rename(molb_full = dosage), by = "name") %>%
  left_join(dosage(skim)    %>% dplyr::rename(skim = dosage),       by = "name") %>%
  left_join(dosage(brb)     %>% dplyr::rename(brbseq = dosage),     by = "name") %>%
  mutate(is_B73_check = name %in% checks) %>% arrange(molb_full)
dose
cat(sprintf("mean donor dosage — MolB-RTI full %.3f | MolB-RTI wsfilt %.3f | Skim-RTI (shared) %.3f | BrB-RTI (shared) %.3f\n",
            mean(dose$molb_full), mean(dose$molb_wsfilt),
            mean(dose$skim[!is.na(dose$skim)]),
            mean(dose$brbseq[!is.na(dose$brbseq)])))
cat(sprintf("B73-bulk checks dosage (MolB-RTI full): %s — expect ≈ 0\n",
            paste(sprintf("%s=%.3f", checks,
                          dose$molb_full[match(checks, dose$name)]), collapse = ", ")))
```

## State on the common grid

```{r state-grid}
paired <- map_dfr(samples, function(s) {
  a <- state_at(mb, s); k <- state_at(skim, s); ok <- !is.na(a) & !is.na(k)
  tibble(name = s, molb = a[ok], skim = k[ok])
})
```

## Confusion matrix — MolB-RTI vs Skim-RTI

Rows = MolB-RTI state (wsfilt, r=8), cols = Skim-RTI (per-taxon r=8); cell = count (row %).

```{r confusion}
cm <- table(molb = factor(paired$molb, 0:2), skim = factor(paired$skim, 0:2))
overall <- sum(diag(cm)) / sum(cm)
prop <- sweep(cm, 1, rowSums(cm), "/")
disp <- matrix(sprintf("%d (%.1f%%)", cm, 100*prop), nrow = 3,
               dimnames = list(`MolB-RTI \\ Skim-RTI` = c("0 recurrent","1 het","2 donor"),
                               c("0 recurrent","1 het","2 donor")))
knitr::kable(disp, align = "r",
             caption = sprintf("MolB-RTI (rows) vs Skim-RTI (cols): count (row %%). Overall concordance %.3f.", overall))
dp <- paired$molb %in% c(1, 2)
cat(sprintf("MolB-RTI donor-present markers: %d | Skim-RTI also donor-present: %.1f%%\n",
            sum(dp), 100*mean(paired$skim[dp] %in% c(1,2))))
cat(sprintf("MolB-RTI homozygous-donor markers: %d | Skim-RTI also hom-donor: %.1f%%\n",
            sum(paired$molb == 2), 100*mean(paired$skim[paired$molb == 2] == 2)))
```

## BrB-RTI cross-check (3′ RNA-seq, shared PN4 NILs)

BrB-RTI state on the common grid for the `r length(brb_samples)` shared samples,
against MolB-RTI (reference) and Skim-RTI. RNA-seq covers only expressed regions at
low depth, so RTIGER sees sparse ALT reads and fragments — treat BrB-RTI as a noisy
qualitative read of donor presence, not a segment-resolution caller.

```{r brb-grid}
paired_b <- map_dfr(brb_samples, function(s) {
  g <- state_at(mb, s); r <- state_at(brb, s); k <- state_at(skim, s)
  ok <- !is.na(g) & !is.na(r)
  tibble(name = s, molb = g[ok], brbseq = r[ok],
         skim = if (s %in% samples) k[ok] else NA_integer_)
})
```

Rows = BrB-RTI state, cols = MolB-RTI (wsfilt, r=8); cell = count (row %).

```{r brb-vs-molb}
cm_bg <- table(brbseq = factor(paired_b$brbseq, 0:2), molb = factor(paired_b$molb, 0:2))
prop  <- sweep(cm_bg, 1, rowSums(cm_bg), "/")
disp  <- matrix(sprintf("%d (%.1f%%)", cm_bg, 100*prop), nrow = 3,
                dimnames = list(`BrB-RTI \\ MolB-RTI` = c("0 recurrent","1 het","2 donor"),
                                c("0 recurrent","1 het","2 donor")))
knitr::kable(disp, align = "r",
             caption = sprintf("BrB-RTI (rows) vs MolB-RTI (cols): count (row %%). Overall concordance %.3f.",
                               sum(diag(cm_bg))/sum(cm_bg)))
dp <- paired_b$molb %in% c(1, 2)
cat(sprintf("MolB-RTI donor-present markers: %d | BrB-RTI also donor-present: %.1f%%\n",
            sum(dp), 100*mean(paired_b$brbseq[dp] %in% c(1,2))))
cat(sprintf("BrB-RTI donor-present markers: %d | MolB-RTI also donor-present: %.1f%%\n",
            sum(paired_b$brbseq %in% c(1,2)),
            100*mean(paired_b$molb[paired_b$brbseq %in% c(1,2)] %in% c(1,2))))
```

Rows = BrB-RTI state, cols = Skim-RTI (per-taxon r=8); cell = count (row %). Restricted
to samples on all three platforms.

```{r brb-vs-skim}
pb3   <- paired_b %>% filter(!is.na(skim))
cm_bk <- table(brbseq = factor(pb3$brbseq, 0:2), skim = factor(pb3$skim, 0:2))
prop  <- sweep(cm_bk, 1, rowSums(cm_bk), "/")
disp  <- matrix(sprintf("%d (%.1f%%)", cm_bk, 100*prop), nrow = 3,
                dimnames = list(`BrB-RTI \\ Skim-RTI` = c("0 recurrent","1 het","2 donor"),
                                c("0 recurrent","1 het","2 donor")))
knitr::kable(disp, align = "r",
             caption = sprintf("BrB-RTI (rows) vs Skim-RTI (cols): count (row %%). Overall concordance %.3f.",
                               sum(diag(cm_bk))/sum(cm_bk)))
```

## Segment scale & crossover count — MolB-RTI vs Skim-RTI vs Null

```{r seg-scale}
metrics <- function(seg, lab, nsamp) {
  allseg <- (seg$end_bp - seg$start_bp + 1)/1e6
  db <- donor_blocks(seg)
  tibble(source = lab,
         donor_blocks_per_sample = round(nrow(db)/nsamp, 1),
         median_all_seg_Mb = round(median(allseg), 1),
         median_donor_block_Mb = round(median(db$size_mb), 2),
         median_COs = median(co_count(seg)$COs))
}
mb_s   <- mb   %>% filter(name %in% samples)
skim_s <- skim %>% filter(name %in% samples)
brb_s  <- brb  %>% filter(name %in% brb_samples)
bind_rows(
  metrics(mb_s,    "MolB-RTI (wsfilt, r=8)", length(samples)),
  metrics(mb_full %>% filter(name %in% samples), "MolB-RTI (full, r=8)", length(samples)),
  metrics(skim_s,  "Skim-RTI (r=8)",        length(samples)),
  metrics(brb_s,   "BrB-RTI (r=8, PN4 only)", length(brb_samples)),
  metrics(sim_seg, "Null (simulation)", n_distinct(sim_seg$name))
)
```

## Introgression-size distribution

```{r intro-sizes}
molb_b <- donor_blocks(mb_s)   %>% mutate(src = "MolB-RTI (wsfilt, r=8)")
skim_b <- donor_blocks(skim_s) %>% mutate(src = "Skim-RTI (r=8)")
brb_b  <- donor_blocks(brb_s)  %>% mutate(src = "BrB-RTI (r=8, PN4)")
sim_b  <- donor_blocks(sim_seg)
ggplot(bind_rows(molb_b, skim_b, brb_b), aes(x = size_mb, fill = src)) +
  geom_histogram(aes(y = after_stat(density)), position = "identity", alpha = 0.5, bins = 30) +
  geom_freqpoly(data = sim_b, aes(x = size_mb, y = after_stat(density)),
                bins = 30, colour = "grey20", linewidth = 0.9, inherit.aes = FALSE) +
  scale_fill_manual(values = c("MolB-RTI (wsfilt, r=8)" = "forestgreen", "Skim-RTI (r=8)" = "purple4",
                               "BrB-RTI (r=8, PN4)" = "orange3")) +
  scale_x_log10() +
  labs(x = "introgression size (Mb, log scale)", y = "density", fill = NULL,
       title = "Introgression-size distribution (2-state, r=8)",
       subtitle = "MolB-RTI/Skim-RTI = all 16 NILs; BrB-RTI = 8 shared PN4; grey line = Null (consensus-map simulation)") +
  theme_bw()
```

## Donor-call composition — het vs homozygous donor

```{r donor-composition}
# per-name marker total differs by source (MolB/Skim on full grid, BrB on its
# shared subset), so each carries its own denominator
tot_gs <- count(paired,   name, name = "tot")
tot_b  <- count(paired_b, name, name = "tot")
comp <- bind_rows(
  paired   %>% transmute(name, platform = "MolB-RTI (wsfilt, r=8)", state = molb) %>%
    filter(state %in% c(1,2)) %>% count(name, platform, state) %>% left_join(tot_gs, by = "name"),
  paired   %>% transmute(name, platform = "Skim-RTI (r=8)",        state = skim) %>%
    filter(state %in% c(1,2)) %>% count(name, platform, state) %>% left_join(tot_gs, by = "name"),
  paired_b %>% transmute(name, platform = "BrB-RTI (r=8, PN4)", state = brbseq) %>%
    filter(state %in% c(1,2)) %>% count(name, platform, state) %>% left_join(tot_b, by = "name")) %>%
  mutate(pct = n/tot, state = factor(state, c(1,2), c("het","homozygous donor")),
         platform = factor(platform, c("MolB-RTI (wsfilt, r=8)", "Skim-RTI (r=8)", "BrB-RTI (r=8, PN4)")))
ord <- paired %>% group_by(name) %>% summarise(d = mean(molb)/2) %>% arrange(d) %>% pull(name)
comp$name <- factor(comp$name, levels = ord)
ggplot(comp, aes(y = name, x = pct, fill = state)) +
  geom_col(position = "stack") + facet_wrap(~ platform) +
  scale_fill_manual(values = c("het" = "forestgreen", "homozygous donor" = "purple4")) +
  scale_x_continuous(labels = scales::percent) +
  labs(x = "% of genome called donor", y = NULL, fill = NULL,
       title = "Donor calls as % of genome (recurrent not drawn)",
       subtitle = "BrB-RTI covers only the 8 shared PN4 samples") + theme_bw()
```

## Chromosome painting — MolB-RTI vs Skim-RTI vs BrB-RTI

Full-genome ancestry painting, three lanes per sample: **top = MolB-RTI (wsfilt,
r=8), middle = Skim-RTI (r=8), bottom = BrB-RTI (r=8)**. Chromosomes are
proportional-width facet columns; segments are painted over their exact
`start_bp→end_bp` extent (`geom_rect`). Palette follows the BzeaSeq convention
(`REF` gold, `HET` springgreen4, `ALT` purple4). Lines are ordered by MolB-RTI donor
dosage with the two B73-bulk checks pinned on top; `LowQuality` lines (sample-map QC
flag) are in **bold**. The BrB-RTI lane is present only for the 8 shared PN4 samples
— the other lines show a blank bottom lane (no BRB-seq counterpart in `run_full`).

```{r chr-painting, fig.width=12, fig.height=16}
pal <- c(REF = "gold", HET = "springgreen4", ALT = "purple4")
as_paint <- function(seg, method) tibble(
  name = seg$name, chr = seg$chr, start_bp = seg$start_bp, end_bp = seg$end_bp,
  state = factor(seg$state, 0:2, c("REF","HET","ALT")), method = method)

dose_ord <- mb %>% mutate(len = end_bp - start_bp + 1) %>% group_by(name) %>%
  summarise(d = sum(len * state/2)/sum(len), .groups = "drop")
ord  <- c(checks, dose_ord %>% filter(!name %in% checks) %>% arrange(d) %>% pull(name))
lowq <- smap$sample_id[smap$quality == "LowQuality" & !is.na(smap$quality)]
name_lab <- function(x) if_else(x %in% lowq, paste0("**", x, "**"), x)

paint3 <- bind_rows(
  as_paint(mb,                                      "MolB-RTI (wsfilt, r=8)"),
  as_paint(skim %>% filter(name %in% unique(mb$name)), "Skim-RTI (r=8)"),
  as_paint(brb,                                     "BrB-RTI (r=8)")) %>%
  mutate(chr    = factor(chr, levels = 1:10, labels = paste0("chr", 1:10)),
         name   = factor(name_lab(name), levels = name_lab(ord)),
         method = factor(method, c("MolB-RTI (wsfilt, r=8)","Skim-RTI (r=8)","BrB-RTI (r=8)")),
         ylo = case_when(method == "MolB-RTI (wsfilt, r=8)" ~ 0.69, method == "Skim-RTI (r=8)" ~ 0.35, TRUE ~ 0.00),
         yhi = case_when(method == "MolB-RTI (wsfilt, r=8)" ~ 1.00, method == "Skim-RTI (r=8)" ~ 0.66, TRUE ~ 0.32))

ggplot(paint3, aes(xmin = start_bp/1e6, xmax = end_bp/1e6,
                   ymin = ylo, ymax = yhi, fill = state)) +
  geom_rect() +
  facet_grid(name ~ chr, scales = "free_x", space = "free_x", switch = "y") +
  scale_fill_manual(values = pal, drop = FALSE) +
  scale_x_continuous(labels = scales::unit_format(unit = "", scale = 1)) +
  labs(title = "Ancestry calls — 16 lines, 3 source–caller combos per sample",
       subtitle = "per sample top→bottom: MolB-RTI (wsfilt) / Skim-RTI / BrB-RTI (all r=8) | gold REF, green HET, purple ALT | bold = LowQuality | blank bottom lane = no BrB-RTI",
       x = "Position (Mb)", fill = "Genotype") +
  ggpubr::theme_classic2(base_size = 9) +
  theme(legend.position = "bottom",
        strip.background = element_blank(),
        strip.text.x = element_text(face = "bold"),
        strip.text.y.left = ggtext::element_markdown(angle = 0, hjust = 1, size = 7),
        panel.spacing.x = unit(0.05, "lines"),
        panel.spacing.y = unit(0.05, "lines"),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        axis.title.y = element_blank(), axis.line.y = element_blank(),
        axis.text.x = element_text(size = 6))
```

**Read:** the BrB-RTI lane (bottom) is sparse and speckled relative to the MolB-RTI
and Skim-RTI lanes — short, scattered donor blocks rather than the long contiguous
introgressions MolB-RTI resolves, and present only for the 8 PN4 lines. It picks up
the largest MolB-RTI donor regions in places but misses most, consistent with the
~13% marker-level donor recall above.

## Findings

- **Real-count pipeline.** Source #5 is target-capture sequencing (~110×), so RTIGER runs on real
  `REF_COUNT`/`ALT_COUNT` like the skim — not the array-style 0/1/2 pseudo-counts. All 16
  samples pass with no QC filter; the data is uniformly high-quality (missing ~0.2%).
- **Polarity & controls check out.** Per-sample donor dosage is low (BC2S3-like, far from
  the ~0.875 of a flipped polarity), and the two B73-bulk checks read ≈ 0 donor — a
  built-in validity test passed without external truth.
- **Cross-source concordance between MolB-RTI and Skim-RTI is high** at marker level;
  per-sample dosage tracks Skim-RTI in rank on the shared NILs, with MolB-RTI reading
  somewhat more donor (especially het) than the coverage-limited Skim-RTI — the expected
  direction (the ~110× target-sequencing platform resolves heterozygosity the ~0.4× skim misses).
- **The simulation is a null, not truth** — consensus-map (~1,783 cM) modeled expectation;
  divergence from it is an observation, not an error. "Resolution" = the detection limit
  ≈ r × marker spacing at r=8, distinct from the (larger) median segment size, a result.
- **BrB-RTI (source #6) is a noisy qualitative layer, not a segment caller.** It is fed to
  RTIGER as real counts like the others, but 3′ RNA-seq covers only expressed regions at
  ~1.5–3× with sparse ALT reads, so the HMM fragments into many short donor blocks and the
  per-sample donor dosage is compressed toward zero (well below the MolB-RTI/Skim-RTI estimate
  and the BC2S3 ≈ 0.125 expectation). On the 8 shared PN4 NILs the headline MolB-RTI concordance
  (~0.80) is almost entirely recurrent-state agreement: BrB-RTI recovers only ~13% of the
  MolB-RTI donor markers, and most of its own donor calls are not confirmed by MolB-RTI. It does
  not resolve het-vs-homozygous donor or block boundaries, and the `PN4_SID326` B73-bulk check
  does **not** read cleanly recurrent (RNA noise over-segments it). So BrB-RTI here is at best
  a coarse hint of where teosinte is, not an independent ancestry map — useful as a sanity
  layer, not a caller.
```