---
title: "MolBreeding GBTS (real counts) vs SNP50K skim — ancestry-call concordance"
author: "Fausto Rodríguez-Zapata"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    df-print: paged
    embed-resources: true
execute:
  warning: false
  message: false
params:
  mb_calls:   "data/molbreeding_gbts/SNP_wsfilt/calls_common_schema.csv"  # GBTS real counts, wideseq-filtered, r=8, all 16
  mb_full:    "data/molbreeding_gbts/SNP/calls_common_schema.csv"         # GBTS real counts, all 20,347 v5 SNPs, r=8
  skim_calls: "data/rtiger_50K/calls_taxa_r8.csv"                         # skim, per-taxon r=8
  sample_map: "data/molbreeding_45k_sample_map.tsv"
  mb_counts:  "data/molbreeding_gbts/SNP_wsfilt/counts"                   # filtered marker grid (RTIGER 6-col)
  sim_seg:    "results/rtiger_45k_benchmark/truth_segments.csv"           # consensus-map simulation (NULL)
  seqlengths: "data/rtiger_50K/seqlengths.csv"
---

**Repo:** [sawers-rellan-labs/zealtiger](https://github.com/sawers-rellan-labs/zealtiger)

Cross-validation of ancestry source **#2 (SNP50K skim, ~0.4×)** against source **#5
(MolBreeding GBTS)** for the BZea NILs genotyped on both platforms.

**This is the real-count pipeline.** MolBreeding source #5 is **GBTS target-capture
sequencing** (~110× depth, ~0.2% missing), *not* a hybridization array — so it is fed to
RTIGER as **real allele counts** (`REF_COUNT`/`ALT_COUNT` from the GATK table), exactly
like the skim, rather than the 0/1/2 pseudo-counts of the earlier array-style attempt.

**Two design choices:**

1. **No line-QC filter.** The array-era "Crappy"/`Good` split was driven by array
   het-rate; with real ~110× counts every sample is clean (missing ~0.2%), so **all 16
   samples are kept** — 14 teosinte NILs + 2 B73-bulk checks. The checks are a built-in
   control (they should read ≈ pure recurrent).
2. **SNP filter (`wsfilt`).** Keep only markers overlapping the **wideseq
   teosinte-vs-B73 panel** (9,157 of the 20,347 lifted v5 SNPs are donor-informative,
   with consistent REF=B73 / ALT=teosinte polarity across sources). The full 20,347-SNP
   fit is shown alongside for reference.

**Rigidity r = 8 for both platforms** — matched to the skim's per-taxon r=8 (Source 2),
so the comparison uses one consistent rigidity rather than ad-hoc per-platform tuning.
Calls are RTIGER segments on B73 NAM v5 (`name, chr, start_bp, end_bp, state`;
0 = recurrent/B73, 1 = het, 2 = homozygous donor).

> The consensus-map BC2S3 **simulation** (source #1) is a **null expectation** under the
> modeled recombination process (MaizeGDB consensus map, ~1,783 cM; `m=10` interference;
> BC2S3 pedigree), **not ground truth**.

## Setup

```{r setup}
library(tidyverse); library(GenomicRanges); library(S4Vectors)
source("R/06_score_rtiger.R")

mb      <- read_csv(params$mb_calls,   show_col_types = FALSE)   # GBTS wsfilt, r=8
mb_full <- read_csv(params$mb_full,    show_col_types = FALSE)   # GBTS full,  r=8
skim    <- read_csv(params$skim_calls, show_col_types = FALSE)   # skim, per-taxon r=8
smap    <- read_tsv(params$sample_map, show_col_types = FALSE)
sl      <- read_csv(params$seqlengths, show_col_types = FALSE)
sim_seg <- read_csv(params$sim_seg,    show_col_types = FALSE)

# common wideseq marker grid, from the filtered RTIGER count files
cf <- read_tsv(list.files(params$mb_counts, full.names = TRUE)[1],
               col_names = c("chr_label","bp","rb","rc","ab","ac"), show_col_types = FALSE)
grid    <- tibble(chr = as.integer(sub("chr","",cf$chr_label)), bp = cf$bp)
spacing <- sum(sl$len) / nrow(grid)

# samples present on both platforms (the 2 B73 checks have no skim NIL match)
samples <- intersect(unique(mb$name), unique(skim$name))
cat(sprintf("MolBreeding GBTS samples: %d (all kept, no QC) | shared with skim: %d\n",
            n_distinct(mb$name), length(samples)))
cat(sprintf("wideseq markers: %d (~%.0f kb spacing; r=8 detection limit ≈ %.1f Mb)\n",
            nrow(grid), spacing/1e3, 8*spacing/1e6))

donor_blocks <- function(seg) seg %>% arrange(name, chr, start_bp) %>% group_by(name, chr) %>%
  mutate(run = cumsum(c(TRUE, diff(state > 0) != 0)), donor = state > 0) %>% filter(donor) %>%
  group_by(name, chr, run) %>% summarise(size_mb = (max(end_bp)-min(start_bp))/1e6, .groups = "drop")
co_count <- function(seg) seg %>% arrange(name, chr, start_bp) %>% group_by(name, chr) %>%
  summarise(tr = sum(abs(diff(as.integer(state > 0)))), .groups = "drop") %>%
  group_by(name) %>% summarise(COs = sum(tr), .groups = "drop")
state_at <- function(df, s) {
  d <- df[df$name == s, ]
  d$state[findOverlaps(GRanges(grid$chr, IRanges(grid$bp, width = 1)),
                       GRanges(d$chr, IRanges(d$start_bp, d$end_bp)), select = "first")]
}
```

## Donor dosage & the B73-bulk control

Per-sample donor dosage (bp-weighted `state/2`). BC2S3 expectation ≈ 0.125; a flipped
REF/ALT polarity would read ≈ 0.875. The two B73-bulk checks (`PN4_SID326`, `PN6_SID576`)
should read ≈ 0 — a built-in validity test that needs no external truth.

```{r dosage}
dosage <- function(seg) seg %>% mutate(len = end_bp - start_bp + 1) %>%
  group_by(name) %>% summarise(dosage = sum(len * state/2) / sum(len), .groups = "drop")
checks <- c("PN4_SID326", "PN6_SID576")
dose <- dosage(mb) %>% dplyr::rename(gbts_wsfilt = dosage) %>%
  left_join(dosage(mb_full) %>% dplyr::rename(gbts_full = dosage), by = "name") %>%
  left_join(dosage(skim)    %>% dplyr::rename(skim = dosage),       by = "name") %>%
  mutate(is_B73_check = name %in% checks) %>% arrange(gbts_full)
dose
cat(sprintf("mean donor dosage — GBTS full %.3f | GBTS wsfilt %.3f | skim (shared) %.3f\n",
            mean(dose$gbts_full), mean(dose$gbts_wsfilt),
            mean(dose$skim[!is.na(dose$skim)])))
cat(sprintf("B73-bulk checks dosage (GBTS full): %s — expect ≈ 0\n",
            paste(sprintf("%s=%.3f", checks,
                          dose$gbts_full[match(checks, dose$name)]), collapse = ", ")))
```

## State on the common grid

```{r state-grid}
paired <- map_dfr(samples, function(s) {
  a <- state_at(mb, s); k <- state_at(skim, s); ok <- !is.na(a) & !is.na(k)
  tibble(name = s, gbts = a[ok], skim = k[ok])
})
```

## Confusion matrix — GBTS vs skim

Rows = GBTS state (wideseq-filtered, r=8), cols = skim (per-taxon r=8); cell = count (row %).

```{r confusion}
cm <- table(gbts = factor(paired$gbts, 0:2), skim = factor(paired$skim, 0:2))
overall <- sum(diag(cm)) / sum(cm)
prop <- sweep(cm, 1, rowSums(cm), "/")
disp <- matrix(sprintf("%d (%.1f%%)", cm, 100*prop), nrow = 3,
               dimnames = list(`gbts \\ skim` = c("0 recurrent","1 het","2 donor"),
                               c("0 recurrent","1 het","2 donor")))
knitr::kable(disp, align = "r",
             caption = sprintf("GBTS (rows) vs skim (cols): count (row %%). Overall concordance %.3f.", overall))
dp <- paired$gbts %in% c(1, 2)
cat(sprintf("GBTS donor-present markers: %d | skim also donor-present: %.1f%%\n",
            sum(dp), 100*mean(paired$skim[dp] %in% c(1,2))))
cat(sprintf("GBTS homozygous-donor markers: %d | skim also hom-donor: %.1f%%\n",
            sum(paired$gbts == 2), 100*mean(paired$skim[paired$gbts == 2] == 2)))
```

## Segment scale & crossover count — GBTS vs skim vs null

```{r seg-scale}
metrics <- function(seg, lab, nsamp) {
  allseg <- (seg$end_bp - seg$start_bp + 1)/1e6
  db <- donor_blocks(seg)
  tibble(source = lab,
         donor_blocks_per_sample = round(nrow(db)/nsamp, 1),
         median_all_seg_Mb = round(median(allseg), 1),
         median_donor_block_Mb = round(median(db$size_mb), 2),
         median_COs = median(co_count(seg)$COs))
}
mb_s   <- mb   %>% filter(name %in% samples)
skim_s <- skim %>% filter(name %in% samples)
bind_rows(
  metrics(mb_s,    "GBTS wsfilt (r=8)", length(samples)),
  metrics(mb_full %>% filter(name %in% samples), "GBTS full (r=8)", length(samples)),
  metrics(skim_s,  "skim (r=8)",        length(samples)),
  metrics(sim_seg, "simulation (null)", n_distinct(sim_seg$name))
)
```

## Introgression-size distribution

```{r intro-sizes}
gbts_b <- donor_blocks(mb_s)   %>% mutate(src = "GBTS wsfilt (r=8)")
skim_b <- donor_blocks(skim_s) %>% mutate(src = "skim (r=8)")
sim_b  <- donor_blocks(sim_seg)
ggplot(bind_rows(gbts_b, skim_b), aes(x = size_mb, fill = src)) +
  geom_histogram(aes(y = after_stat(density)), position = "identity", alpha = 0.5, bins = 30) +
  geom_freqpoly(data = sim_b, aes(x = size_mb, y = after_stat(density)),
                bins = 30, colour = "grey20", linewidth = 0.9, inherit.aes = FALSE) +
  scale_fill_manual(values = c("GBTS wsfilt (r=8)" = "forestgreen", "skim (r=8)" = "purple4")) +
  scale_x_log10() +
  labs(x = "introgression size (Mb, log scale)", y = "density", fill = NULL,
       title = "Introgression-size distribution (all 16 NILs, 2-state, r=8)",
       subtitle = "grey line = consensus-map simulation (null expectation)") + theme_bw()
```

## Donor-call composition — het vs homozygous donor

```{r donor-composition}
comp <- bind_rows(
  paired %>% transmute(name, platform = "GBTS wsfilt (r=8)", state = gbts),
  paired %>% transmute(name, platform = "skim (r=8)",        state = skim)) %>%
  filter(state %in% c(1, 2)) %>% count(name, platform, state) %>%
  left_join(count(paired, name, name = "tot"), by = "name") %>%
  mutate(pct = n/tot, state = factor(state, c(1,2), c("het","homozygous donor")))
ord <- paired %>% group_by(name) %>% summarise(d = mean(gbts)/2) %>% arrange(d) %>% pull(name)
comp$name <- factor(comp$name, levels = ord)
ggplot(comp, aes(y = name, x = pct, fill = state)) +
  geom_col(position = "stack") + facet_wrap(~ platform) +
  scale_fill_manual(values = c("het" = "forestgreen", "homozygous donor" = "purple4")) +
  scale_x_continuous(labels = scales::percent) +
  labs(x = "% of genome called donor", y = NULL, fill = NULL,
       title = "Donor calls as % of genome (recurrent not drawn)") + theme_bw()
```

## Findings

- **Real-count pipeline.** Source #5 is GBTS sequencing (~110×), so RTIGER runs on real
  `REF_COUNT`/`ALT_COUNT` like the skim — not the array-style 0/1/2 pseudo-counts. All 16
  samples pass with no QC filter; the data is uniformly high-quality (missing ~0.2%).
- **Polarity & controls check out.** Per-sample donor dosage is low (BC2S3-like, far from
  the ~0.875 of a flipped polarity), and the two B73-bulk checks read ≈ 0 donor — a
  built-in validity test passed without external truth.
- **Cross-source concordance with the skim is high** at marker level; per-sample dosage
  tracks the skim in rank on the shared NILs, with GBTS reading somewhat more donor
  (especially het) than the coverage-limited skim — the expected direction (the ~110×
  GBTS resolves heterozygosity the ~0.4× skim misses).
- **The simulation is a null, not truth** — consensus-map (~1,783 cM) modeled expectation;
  divergence from it is an observation, not an error. "Resolution" = the detection limit
  ≈ r × marker spacing at r=8, distinct from the (larger) median segment size, a result.
```