---
title: "Sample-identity check — the target sequencing PN4_SID330 mislabel"
author: "Fausto Rodríguez-Zapata"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    df-print: paged
    embed-resources: true
execute:
  warning: false
  message: false
params:
  mb_calls:   "results/sim_calibration/molb_calls/SNP_wsfilt/calls_common_schema.csv"  # target-sequencing wsfilt, r=2
  mb_counts:  "results/sim_calibration/molb_calls/SNP_wsfilt/counts"                    # wsfilt marker grid (9,157)
  skim_calls: "data/rtiger_50K/calls_taxa_r5.csv"                                       # all 1393 skim RTIGER r=5
  query:      "PN4_SID330"
  seed:       1
---

**Repo:** [sawers-rellan-labs/zealtiger](https://github.com/sawers-rellan-labs/zealtiger)

While painting target sequencing vs skim ancestry calls for the 16 dual-platform NILs
(`molbreeding_vs_skim_calibrated.qmd`), **`PN4_SID330`** was the one line whose two
platforms disagreed wildly — not in amount but in *pattern*. This notebook tests
whether the two `PN4_SID330` calls even come from the same DNA, and if not, finds
the true identity by scanning **all 1393 skim samples**.

**Result up front:** the target sequencing sample labelled `PN4_SID330` is **PN4_SID322** — a
label/well swap on the target sequencing (MolBreeding) side. The skim `PN4_SID330` is a
distinct, correctly-labelled individual.

## Setup

```{r setup}
library(tidyverse); library(data.table); library(GenomicRanges)
set.seed(params$seed)

mb   <- read_csv(params$mb_calls,   show_col_types = FALSE)   # target-sequencing wsfilt, r=2 (16 samples)
skim <- read_csv(params$skim_calls, show_col_types = FALSE)   # skim RTIGER r=5 (1393 samples)
cf   <- read_tsv(list.files(params$mb_counts, full.names = TRUE)[1],
                 col_names = c("chr_label","bp","rb","rc","ab","ac"), show_col_types = FALSE)
grid <- tibble(chr = as.integer(sub("chr","",cf$chr_label)), bp = cf$bp) %>% mutate(gi = row_number())
gg   <- GRanges(grid$chr, IRanges(grid$bp, width = 1))

# state-on-grid (0/1/2) for one sample's segment table
state_on_grid <- function(seg) {
  h <- findOverlaps(gg, GRanges(seg$chr, IRanges(seg$start_bp, seg$end_bp)), select = "first")
  seg$state[h]
}
# identity metrics: donor-present Jaccard (robust to skim het-undercall) + state concordance
jac  <- function(a, b) { ok <- !is.na(a) & !is.na(b); a <- a[ok] > 0; b <- b[ok] > 0
  if (sum(a | b) == 0) return(NA_real_); sum(a & b) / sum(a | b) }
conc <- function(a, b) { ok <- !is.na(a) & !is.na(b); if (!any(ok)) return(NA_real_); mean(a[ok] == b[ok]) }

cat(sprintf("query = target sequencing-%s | skim panel = %d samples | grid = %d sites\n",
            params$query, n_distinct(skim$name), nrow(grid)))
```

## Identity scan — target-sequencing PN4_SID330 vs all 1393 skim samples

The query is the target sequencing `PN4_SID330` donor pattern; we score it against every skim
sample's calls on the common grid. A correctly-labelled sample matches its own
label best (donor Jaccard ≈ 0.65–0.98, as the other 13 dual-platform NILs do).

```{r scan}
q  <- state_on_grid(mb %>% filter(name == params$query))   # target sequencing query
qd <- q > 0

# all skim samples on the grid in one pass (data.table foverlaps)
sk  <- as.data.table(skim)[, .(name, chr = as.character(chr), start_bp, end_bp, state)]
gdt <- as.data.table(grid)[, .(gi, chr = as.character(chr), start = bp, end = bp)]
setkey(sk, chr, start_bp, end_bp)
ov <- foverlaps(gdt, sk, by.x = c("chr","start","end"),
                by.y = c("chr","start_bp","end_bp"), type = "within", nomatch = NULL)

res <- ov[, {
  v <- rep(NA_integer_, nrow(grid)); v[gi] <- state
  .(donorJ = jac(q, v), concord = conc(q, v), skim_donor = mean(v > 0, na.rm = TRUE))
}, by = name][order(-donorJ)]

knitr::kable(head(res, 15), digits = 3,
             caption = "Top 15 skim matches to target-sequencing PN4_SID330 (by donor Jaccard)")

own <- which(res$name == params$query)
cat(sprintf("\nlabelled skim %s: rank %d / %d  donorJ = %.3f  concord = %.3f\n",
            params$query, own, nrow(res), res$donorJ[own], res$concord[own]))
```

**Read:** a single decisive hit — **PN4_SID322** at donor Jaccard **0.84** /
concordance **0.91** — far above the runner-up (~0.47) and the line's own label
(**0.10, rank ~506/1393**). target sequencing-`PN4_SID330` is the same DNA as skim-`PN4_SID322`.

## Ruling out "they're just sibs"

Both `PN4_SID322` and `PN4_SID330` are family **Zx.0030**, so the match could in
principle be close relatives rather than identical DNA. The test: compare the two
**skim** samples directly. If they were near-identical sibs, the match would be
ambiguous; if they are distinct, target-sequencing 330 ≡ skim-322 can only be a swap.

```{r sibs}
g330 <- state_on_grid(mb   %>% filter(name == "PN4_SID330"))
k322 <- state_on_grid(skim %>% filter(name == "PN4_SID322"))
k330 <- state_on_grid(skim %>% filter(name == "PN4_SID330"))
tibble(
  comparison = c("skim-322 vs skim-330 (both Zx.0030)",
                 "target-sequencing 330 vs skim-322 (the match)",
                 "target-sequencing 330 vs skim-330 (its own label)"),
  donor_Jaccard = c(jac(k322, k330), jac(g330, k322), jac(g330, k330)),
  concordance   = c(conc(k322, k330), conc(g330, k322), conc(g330, k330)),
  donor_frac_A_vs_B = c(sprintf("%.3f / %.3f", mean(k322>0,na.rm=T), mean(k330>0,na.rm=T)),
                        sprintf("%.3f / %.3f", mean(g330>0,na.rm=T), mean(k322>0,na.rm=T)),
                        sprintf("%.3f / %.3f", mean(g330>0,na.rm=T), mean(k330>0,na.rm=T)))) %>%
  knitr::kable(digits = 3)
```

**Read:** skim-322 vs skim-330 is only **J=0.10** — they are *distinct* individuals
(same donor source, independent recombination), no more alike than random NILs. So
target-sequencing 330 matching skim-322 at 0.84 while sharing nothing with skim-330 is **not**
relatedness — it is a sample swap. Donor fractions agree on the match
(target-sequencing 330 0.32 ≈ skim-322 0.30) and disagree on the label (skim-330 0.11).

## Chromosome painting — the swap, in colour

Top = target sequencing `PN4_SID330` (labelled); middle = skim `PN4_SID322` (the J=0.84 match);
bottom = skim `PN4_SID330` (its own label, J=0.10). Palette: REF gold, HET green,
ALT purple.

```{r painting, fig.width=13, fig.height=4.2}
pal <- c(REF = "gold", HET = "springgreen4", ALT = "purple4")
lane <- function(seg, nm, lab) seg %>% filter(name == nm) %>%
  transmute(chr, start_bp, end_bp, state = factor(state, 0:2, c("REF","HET","ALT")), lane = lab)
paint <- bind_rows(
  lane(mb,   "PN4_SID330", "target sequencing  PN4_SID330  (labelled)"),
  lane(skim, "PN4_SID322", "skim  PN4_SID322  (J=0.84 match)"),
  lane(skim, "PN4_SID330", "skim  PN4_SID330  (own label, J=0.10)")) %>%
  mutate(chr  = factor(chr, levels = 1:10, labels = paste0("chr", 1:10)),
         lane = factor(lane, levels = c("target sequencing  PN4_SID330  (labelled)",
                                        "skim  PN4_SID322  (J=0.84 match)",
                                        "skim  PN4_SID330  (own label, J=0.10)")))
ggplot(paint, aes(xmin = start_bp/1e6, xmax = end_bp/1e6, ymin = 0, ymax = 1, fill = state)) +
  geom_rect() +
  facet_grid(lane ~ chr, scales = "free_x", space = "free_x", switch = "y") +
  scale_fill_manual(values = pal, drop = FALSE) +
  scale_x_continuous(labels = scales::unit_format(unit = "", scale = 1)) +
  labs(title = "PN4_SID330 identity — target-sequencing 330 matches skim-322, not skim-330",
       subtitle = "gold REF · green HET · purple ALT", x = "Position (Mb)", fill = "Genotype") +
  ggpubr::theme_classic2(base_size = 11) +
  theme(legend.position = "bottom", strip.background = element_blank(),
        strip.text.y.left = element_text(angle = 0, hjust = 1),
        panel.spacing.x = unit(0.05, "lines"), panel.spacing.y = unit(0.2, "lines"),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        axis.title.y = element_blank(), axis.line.y = element_blank(),
        axis.text.x = element_text(size = 6))
```

**Read:** the target-sequencing 330 and skim-322 lanes share the same introgressions in the same
places — the chr4 ALT block, the chr6 and chr9 het blocks, the chr3 het/ALT region.
The skim-330 lane (its own label) carries entirely different blocks (chr2, chr3
only) and shares essentially nothing with target-sequencing 330.

## Conclusion

- The target sequencing (MolBreeding) sample labelled **`PN4_SID330`** (well **4B6**) is in fact
  **`PN4_SID322`**: donor Jaccard 0.84 / concordance 0.91 vs skim-322, vs 0.10 vs
  its own label, decisive over 1393 candidates.
- It is a **swap**, not relatedness: skim-322 and skim-330 are distinct individuals
  (J=0.10) despite the shared Zx.0030 source.
- The skim `PN4_SID330` is correctly labelled (a distinct low-donor individual);
  the error is on the target sequencing side.
- **Action:** exclude or relabel target sequencing `PN4_SID330` in any per-sample target-sequencing-vs-skim
  comparison. Confirming *how* the swap happened (e.g. a neighbouring-well pipetting
  slip) needs the master DNA-plate layout placing `PN4_SID322`, which is not in the
  available metadata — the identity itself is certain regardless.
```