---
title: "RTIGER vs wideseq — ancestry-calling performance on simulated BC2S3 NILs"
author: "Fausto Rodríguez-Zapata"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    df-print: paged
    embed-resources: true
execute:
  warning: false
  message: false
---

**Repo:** [sawers-rellan-labs/zealtiger](https://github.com/sawers-rellan-labs/zealtiger)

## Scope

Head-to-head benchmark of the two ancestry callers against the **same simulated
ground truth**, each in its native platform/regime (shared `simcross` BC2S3 mosaics,
seed `20260609`, n = 100):

- **RTIGER** (`faustovrz/RTIGER` fork), objective autotune **r = 8**, on the **SNP50K
  skim** (λ ≈ 0.43×, ~50K markers), scored per marker.
- **wideseq** (method B; BzeaSeq `call_ancestry.Rmd`), the recommended **`Kgmm_HMM`**
  3-state caller, on the **teosinte panel** (~27.6M sites, ~12,950/Mb, λ ≈ 0.59× →
  1 Mb bins), scored per bin.

Both are scored against their own truth at native resolution
(`make_rtiger_benchmark.R` + `make_wideseq_benchmark.R` → `score_rtiger.R` /
`score_wideseq.R`). Because the underlying recombination mosaics are identical, the
comparison is fair on truth; differences reflect platform + resolution + method
together — the realistic operational comparison (the two are inherently
different-resolution callers, so there is no single shared input).

> **State coding** (both): 0 = homozygous recurrent (B73), 1 = het, 2 = homozygous
> donor (teosinte). An *introgression* = a maximal donor-present run (state ≥ 1).

## Genotyping accuracy

```{r perf-tables, echo=FALSE}
# Static: scored on the n=100 benchmark (RTIGER autotune r=8; wideseq Kgmm_HMM).
# Reproduce: Rscript score_rtiger.R --rtiger results/rtiger_benchmark/rtiger_out_auto/rtiger_result.rds
#            Rscript score_wideseq.R   (results/{rtiger,wideseq}_benchmark/scoring/)
marker <- tibble::tribble(
  ~state,            ~RTIGER_recall, ~RTIGER_prec, ~wideseq_recall, ~wideseq_prec,
  "0 recurrent",            1.000,        0.999,         0.998,         0.996,
  "1 het",                  0.974,        0.983,         0.971,         0.939,
  "2 donor",                0.996,        0.997,         0.963,         0.995
)
segment <- tibble::tribble(
  ~class,              ~RTIGER_P, ~RTIGER_R, ~RTIGER_F1, ~wideseq_P, ~wideseq_R, ~wideseq_F1,
  "homozygous donor",      0.999,     0.887,      0.939,      0.997,      0.792,      0.882,
  "heterozygous",          0.895,     0.773,      0.830,      0.958,      0.724,      0.825,
  "donor present",         0.980,     0.861,      0.917,      0.989,      0.772,      0.867
)
crossovers <- tibble::tribble(
  ~method,               ~called, ~true, ~co_bias,
  "RTIGER (r=8)",           19.1,  21.8,    -2.71,
  "wideseq (Kgmm_HMM)",     15.4,  19.9,    -4.48
)
knitr::kable(marker, caption = paste(
  "Marker/bin-level recall & precision.",
  "Overall accuracy: RTIGER 0.9986 (0.35% uncovered), wideseq 0.9937 (0% uncovered)."))
knitr::kable(segment, caption = paste(
  "Segment-level (reciprocal overlap >= 0.5). Boundary error (donor-present):",
  "RTIGER ~85 kb; wideseq 0 kb is a 1 Mb-grid artifact (true floor +/-1 Mb)."))
knitr::kable(crossovers, caption = paste(
  "Crossovers per genome vs each method's OWN truth (RTIGER truth has more",
  "detectable COs at 50K-marker resolution) -> compare bias, not absolute counts."))
```

**Takeaways.** Both genotype near-perfectly at the marker/bin level (≥ 99%). The
split is in resolution vs cleanliness:

- **RTIGER recovers more and finer** — higher segment recall everywhere
  (donor-present F1 0.917 vs 0.867) and a smaller crossover deficit (−2.7 vs −4.5),
  because at marker resolution it detects sub-Mb introgressions that wideseq's 1 Mb
  binning merges or misses. Boundaries land within ~85 kb (≈2 marker spacings).
- **wideseq is cleaner / more precise** — higher donor-present precision (0.989 vs
  0.980) and markedly higher het precision (0.958 vs 0.895): the dense per-bin
  aggregation gives confident, low-false-positive calls, but it cannot localize below
  ~1 Mb (the "0 kb" boundary error is a scoring artifact of calls and truth sharing
  bin edges).

## Introgression-size distribution

```{r intro-sizes, echo=FALSE, fig.width=8, fig.height=5}
library(dplyr); library(ggplot2)
blocks <- readr::read_csv("docs/rtiger_vs_wideseq_blocks.csv", show_col_types = FALSE)

summ <- blocks %>% group_by(src) %>%
  summarise(blocks_per_sample = round(n() / dplyr::n_distinct(name), 1),
            median_Mb = round(median(size_mb), 1),
            mean_Mb   = round(mean(size_mb), 1),
            `p10_Mb`  = round(quantile(size_mb, .10), 1),
            `p90_Mb`  = round(quantile(size_mb, .90), 1),
            `pct_under_1Mb` = round(100 * mean(size_mb < 1), 1),
            donor_Mb_per_sample = round(sum(size_mb) / dplyr::n_distinct(name), 0),
            .groups = "drop")
knitr::kable(summ, caption = "Introgression-size summary (donor-present blocks, n=100).")

tb <- filter(blocks, src == "simulation (truth)")
mp <- filter(blocks, src != "simulation (truth)")
ggplot(mp, aes(x = size_mb, fill = src)) +
  geom_histogram(aes(y = after_stat(density)), position = "identity",
                 alpha = 0.5, bins = 30) +
  geom_freqpoly(data = tb, aes(x = size_mb, y = after_stat(density)),
                bins = 30, colour = "grey20", linewidth = 0.9, inherit.aes = FALSE) +
  scale_fill_manual(values = c("RTIGER (r=8)" = "forestgreen",
                               "wideseq (Kgmm_HMM)" = "purple4")) +
  scale_x_log10() +
  labs(x = "introgression size (Mb, log scale)", y = "density", fill = NULL,
       title = "Introgression-size distribution: RTIGER vs wideseq",
       subtitle = "grey line = simulation truth; shared BC2S3 mosaics (n=100)") +
  theme_bw()
```

**Takeaways.**

- **Both preserve total donor burden** (~290 Mb/sample vs truth 294) — the genome-wide
  donor fraction is right either way.
- **RTIGER reproduces the truth distribution**, including the small-block tail: nearly
  the right block count (9.8 vs 10.6/sample), matching median (12.5 vs 11.2 Mb), and it
  resolves down to ~1 Mb. The green histogram sits on the truth line.
- **wideseq is shifted to larger blocks**: fewer blocks (8.0/sample), inflated median
  (17 vs 11 Mb), and **0% below 1 Mb by construction** — the 1 Mb floor erases sub-Mb
  introgressions and merges adjacent ones, repackaging the same donor content into
  fewer, larger blocks.

## Conclusion

The two callers are **complementary**. For the project's **introgression-size
distribution** deliverable, **RTIGER is the faithful instrument** — its size
distribution matches ground truth including the fine (<5 Mb) tail. **wideseq** nails
per-bin genotype and total donor fraction with very clean (high-precision) calls, but
its size distribution is biased toward larger blocks and is blind below ~1 Mb; treat it
as a coarse view (lower bound on block count, upper bound on block size). Use RTIGER
where fine-scale recombination/size matters; use wideseq for clean high-confidence broad
ancestry on the dense teosinte panel.
