---
title: "RTIGER vs wideseq — ancestry-calling performance on simulated BC2S3 NILs"
author: "Fausto Rodríguez-Zapata"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    df-print: paged
    embed-resources: true
execute:
  warning: false
  message: false
---

**Repo:** [sawers-rellan-labs/zealtiger](https://github.com/sawers-rellan-labs/zealtiger)

## Scope

Head-to-head benchmark of the two ancestry callers against **one canonical ground
truth**. The genome simulation is **decoupled from the calling method**
(`make_joint_benchmark.R`): per sample we draw the BC2S3 mosaic **once** and derive
from that single mosaic both platforms' observations and a high-resolution canonical
truth, then score **both callers against that same truth** (n = 100).

- **Canonical truth:** per-marker dosage on a ~5 kb grid (200 markers/Mb) → exact
  segments and crossover count, independent of any caller's resolution.
- **RTIGER** (`faustovrz/RTIGER` fork), rigidity **r = 8** (the objective autotune
  value), on the **SNP50K skim** observation (λ ≈ 0.43×, ~50K markers).
- **wideseq** (method B; BzeaSeq `call_ancestry.Rmd`), recommended **`Kgmm_HMM`**
  caller, on the **teosinte-panel** observation (~12,950 sites/Mb, λ ≈ 0.59× → 1 Mb
  bins).

Both callers' segments are scored with the same `R/06` functions
(`assign_called_state` / `score_markers` / `score_segments` / `score_cos`) against the
same `truth_markers` / `truth_segments`, so **segment recall and crossover counts
share one denominator**. Coverage differs by platform (a property of the assay, not the
genome); resolution differs by method. This supersedes an earlier version that scored
each caller against its own, differently-resolved truth — which flattered the
coarser caller.

> **State coding** (both): 0 = homozygous recurrent (B73), 1 = het, 2 = homozygous
> donor (teosinte). An *introgression* = a maximal donor-present run (state ≥ 1).

## Genotyping accuracy

```{r perf-tables, echo=FALSE}
# Static: score_joint.R on the n=100 joint benchmark (one canonical truth).
# Reproduce: Rscript make_joint_benchmark.R 100;
#            JULIA_HOME=... Rscript fit_rtiger.R 0 8 0 4 results/joint_benchmark/rtiger;
#            Rscript score_joint.R   (-> results/joint_benchmark/scoring/)
marker <- tibble::tribble(
  ~state,        ~RTIGER_recall, ~RTIGER_prec, ~wideseq_recall, ~wideseq_prec,
  "0 recurrent",        1.000,        0.999,         0.998,         0.996,
  "1 het",              0.978,        0.983,         0.962,         0.935,
  "2 donor",            0.997,        0.998,         0.963,         0.994
)
segment <- tibble::tribble(
  ~class,              ~RTIGER_P, ~RTIGER_R, ~RTIGER_F1, ~wideseq_P, ~wideseq_R, ~wideseq_F1,
  "homozygous donor",      0.996,     0.900,      0.946,      0.997,      0.761,      0.863,
  "heterozygous",          0.892,     0.758,      0.820,      0.918,      0.573,      0.705,
  "donor present",         0.978,     0.866,      0.919,      0.979,      0.701,      0.817
)
crossovers <- tibble::tribble(
  ~method,               ~called, ~true, ~co_bias,
  "RTIGER (r=8)",           18.4,  20.9,    -2.57,
  "wideseq (Kgmm_HMM)",     14.8,  20.9,    -6.11
)
knitr::kable(marker, caption = paste(
  "Marker-level recall & precision vs the canonical ~5 kb truth.",
  "Overall accuracy: RTIGER 0.9989 (0.38% uncovered), wideseq 0.9938 (0% uncovered)."))
knitr::kable(segment, caption = paste(
  "Segment-level (reciprocal overlap >= 0.5) vs the SAME truth -> identical n_truth",
  "(874 / 447 / 1321 for the three classes)."))
knitr::kable(crossovers, caption = paste(
  "Crossovers per genome vs the SAME canonical truth (true = 20.9 for both)."))
```

**Takeaways** (now that both are judged against one honest truth):

- **Genotyping is near-perfect for both** at the marker level (RTIGER 0.999, wideseq
  0.994); the split is in *segment recovery*.
- **RTIGER recovers substantially more.** Against the shared truth its segment recall
  leads everywhere — donor-present 0.866 vs 0.701, het 0.758 vs 0.573 — and its
  crossover deficit is far smaller (−2.6 vs −6.1 against the same true 20.9). At ~5 kb /
  50K-marker resolution it detects sub-Mb introgressions that wideseq's 1 Mb binning
  merges or misses.
- **Precision is comparable** (donor-present ~0.98 for both); wideseq keeps a small het
  precision edge (0.918 vs 0.892) but it no longer compensates for the recall gap once
  the truth is shared. (The earlier per-method-truth comparison hid wideseq's misses by
  scoring it against a truth already coarsened to 1 Mb.)

## Introgression-size distribution

```{r intro-sizes, echo=FALSE, fig.width=8, fig.height=5}
library(dplyr); library(ggplot2)
blocks <- readr::read_csv("docs/rtiger_vs_wideseq_blocks.csv", show_col_types = FALSE)

summ <- blocks %>% group_by(src) %>%
  summarise(blocks_per_sample = round(n() / dplyr::n_distinct(name), 1),
            median_Mb = round(median(size_mb), 1),
            mean_Mb   = round(mean(size_mb), 1),
            `p10_Mb`  = round(quantile(size_mb, .10), 1),
            `p90_Mb`  = round(quantile(size_mb, .90), 1),
            `pct_under_1Mb` = round(100 * mean(size_mb < 1), 1),
            donor_Mb_per_sample = round(sum(size_mb) / dplyr::n_distinct(name), 0),
            .groups = "drop")
knitr::kable(summ, caption = "Introgression-size summary (donor-present blocks, n=100).")

tb <- filter(blocks, src == "simulation (truth)")
mp <- filter(blocks, src != "simulation (truth)")
ggplot(mp, aes(x = size_mb, fill = src)) +
  geom_histogram(aes(y = after_stat(density)), position = "identity",
                 alpha = 0.5, bins = 30) +
  geom_freqpoly(data = tb, aes(x = size_mb, y = after_stat(density)),
                bins = 30, colour = "grey20", linewidth = 0.9, inherit.aes = FALSE) +
  scale_fill_manual(values = c("RTIGER (r=8)" = "forestgreen",
                               "wideseq (Kgmm_HMM)" = "purple4")) +
  scale_x_log10() +
  labs(x = "introgression size (Mb, log scale)", y = "density", fill = NULL,
       title = "Introgression-size distribution: RTIGER vs wideseq",
       subtitle = "grey line = simulation truth; ONE genome simulation, n=100") +
  theme_bw()
```

**Takeaways.**

- **Both preserve total donor burden** (~265 Mb/sample vs truth 268) — genome-wide donor
  fraction is right either way.
- **RTIGER reproduces the truth distribution**, including the small-block tail: nearly
  the right block count (9.5 vs 10.4/sample), matching median (11.6 vs 10.2 Mb), and it
  resolves down to ~1 Mb (3.6% < 1 Mb vs truth's 7.3%). Green sits on the truth line.
- **wideseq is shifted to larger blocks**: fewer blocks (8.0/sample), inflated median
  (16 vs 10 Mb), and **0% below 1 Mb by construction** — the 1 Mb floor erases sub-Mb
  introgressions and merges adjacent ones, repackaging the same donor content into
  fewer, larger blocks.

## Conclusion

Judged against a **single canonical truth**, the two callers are complementary but the
ranking on recovery is clear. For the project's **introgression-size distribution**
deliverable, **RTIGER is the faithful instrument** — its size distribution matches
ground truth including the sub-5 Mb tail, and it recovers segments and crossovers far
more completely. **wideseq** is highly precise per call and nails total donor fraction
and per-bin genotype, but its 1 Mb resolution costs real recall (donor-present 0.70,
het 0.57) and crossover recovery (−6.1), and it is blind below ~1 Mb. Use RTIGER where
fine-scale recombination / introgression size matters; use wideseq for clean
high-confidence broad ancestry on the dense teosinte panel.
