---
title: "Cross-experiment sample correspondence — Skim · MolBreeding · BRB-seq"
subtitle: "Which sample id in one experiment is the same biological NIL/check in another"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    df-print: kable
params:
  master:   "data/sample_metadata_master.csv"              # skim/MolBreeding registry (sample -> pedigree, with .B)
  brb_meta: "data/brbseq_metadata_master.csv"              # BRB-seq (sample_id IS the fastq prefix)
  mb_map:   "data/molbreeding_45k_sample_map.tsv"          # MolBreeding target-seq plate map (16 samples)
  skim_seq: "data/skim_sequenced_ids.txt"                 # skim WGS BAMs (PN#_SID#, one/line) = samples WITH sequence data
  brb_seq:  "results/sim_calibration/brbseq_coverage.csv"  # BRB samples WITH sequence data
  outdir:   "results/sample_correspondence"
---

## Purpose

Curate, for **every sample that has sequence data**, the correspondence between the
per-experiment `PN#_SID#` fastq sample id and the **true biological identity** — so we
compare the right genome tracks across experiments.

### The trap this fixes

All three experiments label fastqs with a `PN#_SID#` prefix
(`PN{ceil(SID/96)}_SID{SID}`), but the SID numbering is **per-experiment**, so the *same
prefix is a different biological sample across experiments*. Concretely:

- **Skim** (WGS low-coverage) and **MolBreeding** (target sequencing) **share one id space** —
  same `PN#_SID#` ⇒ same sample. (Skim-RTI on the SNP50K filter and Skim-BIN on the 27M
  wideseq filter are the *same* skim reads.)
- **BRB-seq** (3′ RNA) has its **own** `PN#_SID#` space. Assuming it matched Skim/MolB by
  prefix is what previously compared incomparable samples.

**Cross-experiment identity key = the pedigree string** (drop the trailing `.B` the
Skim/MolB registry carries; BRB pedigrees have none). The `PN#_SID#` prefix is *not* a
cross-experiment key.

### Scope and rules

- **In scope:** the **bzea/ZEAL population** (`project == "bzea"`) — teosinte-introgression
  **NILs** + the **B73** and **Purple** checks. The `landb`/`lanteo` landrace lines are
  excluded (only a handful were sequenced anyway).
- **Sequence-data only:** a sample appears only if it has reads — **skim** from the WGS BAM
  list (`bam_list_1440`, includes the checks), **BRB** from its coverage table. Metadata-only
  and out-of-scope rows are excluded.
- **Label:** NILs → the pedigree string; checks → `B73_<prefix>` / `Purple_<prefix>`
  (each check is unique because checks cannot yet be matched across experiments — see
  [Checks](#checks-to-resolve)).
- Sample class is read from `Genotype`/`sample_group`/`accession`, **never** the `check`
  column (it is unreliable — e.g. BRB "Purple Check" rows carry `check="no"`).

```{r setup}
#| message: false
#| warning: false
library(readr); library(dplyr); library(stringr); library(tidyr); library(knitr)

chop <- function(x) sub("\\.B$", "", x)   # drop the trailing ".B"

## ---- samples WITH sequence data ----
skim_seq <- readLines(params$skim_seq)                              # skim WGS BAMs (PN#_SID#)
brb_seq  <- read_csv(params$brb_seq, show_col_types = FALSE)$name   # BRB sequenced

## ---- Skim / MolBreeding id space (shared) ----
master <- read_csv(params$master, show_col_types = FALSE) %>%
  mutate(pedigree = ifelse(pedigree %in% c("NA", ""), NA, pedigree),
         ped      = ifelse(!is.na(pedigree), chop(pedigree), NA),
         chk_grp  = ifelse(sample_group %in% c("B73", "Purple"), sample_group, NA))
## scope = the bzea/ZEAL population (project == "bzea"); landb/lanteo landrace lines are excluded.
skim_nil <- master %>% filter(project == "bzea", !is.na(ped),     sample %in% skim_seq) %>% transmute(skim_prefix = sample, pedigree = ped)
skim_chk <- master %>% filter(project == "bzea", !is.na(chk_grp), sample %in% skim_seq) %>% transmute(skim_prefix = sample, grp = chk_grp)

## ---- BRB-seq id space (own) ----
brbm <- read_csv(params$brb_meta, show_col_types = FALSE) %>%
  mutate(brb_prefix = sample_id,                              # already PN#_SID#
         is_nil  = str_detect(Genotype, "_P[0-9]"),           # real pedigree -> NIL
         chk_grp = case_when(Genotype == "B73" | accession == "B73"             ~ "B73",
                             accession == "Purple" | str_detect(Genotype, "Purple") ~ "Purple",
                             TRUE ~ NA_character_),
         ped     = ifelse(is_nil, chop(Genotype), NA))
brb_nil <- brbm %>% filter(is_nil,          brb_prefix %in% brb_seq) %>% transmute(brb_prefix, pedigree = ped)
brb_chk <- brbm %>% filter(!is.na(chk_grp), brb_prefix %in% brb_seq) %>% transmute(brb_prefix, grp = chk_grp)

## ---- MolBreeding target-seq (the 16 map rows ARE the sequenced set) ----
molb <- read_tsv(params$mb_map, show_col_types = FALSE) %>%
  transmute(prefix = sample_id, molb_geno = genotype, molb_ped = chop(genotype))

agg <- function(df, key, val) df %>% group_by(.data[[key]]) %>%
  summarise("{val}" := paste(sort(unique(.data[[val]])), collapse = ";"),
            "n_{val}" := n_distinct(.data[[val]]), .groups = "drop")
```

## File 1 — Skim ↔ BRB-seq (matched by pedigree)

```{r file1}
sk <- agg(skim_nil, "pedigree", "skim_prefix")
br <- agg(brb_nil,  "pedigree", "brb_prefix") %>% rename(brbseq_prefix = brb_prefix, n_brbseq_prefix = n_brb_prefix)

nil1 <- full_join(sk, br, by = "pedigree") %>%
  transmute(type = "NIL", label = pedigree, pedigree,
            skim_prefix, brbseq_prefix,
            in_skim = !is.na(skim_prefix), in_brbseq = !is.na(brbseq_prefix)) %>%
  arrange(desc(in_skim & in_brbseq), pedigree)

chk1 <- bind_rows(
  skim_chk %>% transmute(type = grp, label = paste0(grp, "_", skim_prefix), pedigree = NA_character_,
                         skim_prefix, brbseq_prefix = NA_character_, in_skim = TRUE,  in_brbseq = FALSE),
  brb_chk  %>% transmute(type = grp, label = paste0(grp, "_", brb_prefix), pedigree = NA_character_,
                         skim_prefix = NA_character_, brbseq_prefix = brb_prefix, in_skim = FALSE, in_brbseq = TRUE))

file1 <- bind_rows(nil1, chk1)
dir.create(params$outdir, showWarnings = FALSE, recursive = TRUE)
write_csv(file1, file.path(params$outdir, "skim_brbseq_correspondence.csv"))

nil1 %>% count(`NIL status` = case_when(in_skim & in_brbseq ~ "comparable (both sequenced)",
                                        in_skim             ~ "skim-only",
                                        TRUE                ~ "BRB-only")) %>%
  kable(caption = "Skim ↔ BRB-seq NILs by availability")
```

**`r sum(nil1$in_skim & nil1$in_brbseq)` NILs are directly comparable** (sequenced on both
platforms, matched by pedigree). The full per-NIL table is **not embedded here** — it ships
as the CSV for Hannah: `results/sample_correspondence/skim_brbseq_correspondence.csv`
(NILs + checks, with `in_skim`/`in_brbseq` flags).

## File 2 — MolBreeding 3-way (MolB · Skim · BRB-seq)

MolBreeding sequenced only **16 samples**, so rather than a separate Skim↔MolBreeding file we
give the **full 3-way correspondence for those 16** — their MolB id, the shared Skim id, and
(where the same NIL was also sequenced in BRB-seq) the BRB-seq id. MolBreeding and Skim share
one id space (same `PN#_SID#`); the BRB-seq link is by **pedigree** and exists for only a few
of the lines (MolBreeding and BRB-seq otherwise share no id space and overlap on just those
NILs — hence the wide Skim↔BRB correspondence stays in File 1).

```{r file2}
brb_by_ped <- brb_nil %>% group_by(pedigree) %>%
  summarise(brbseq_prefix = paste(sort(unique(brb_prefix)), collapse = ";"), .groups = "drop")

file2 <- molb %>%
  left_join(master %>% transmute(prefix = sample, skim_ped = ped, skim_grp = chk_grp), by = "prefix") %>%
  mutate(grp   = case_when(str_detect(molb_geno, "B73")    ~ "B73",
                           str_detect(molb_geno, "Purple") ~ "Purple",
                           !is.na(skim_grp)                ~ skim_grp,
                           TRUE ~ NA_character_),
         type  = ifelse(!is.na(grp), grp, "NIL"),
         label = ifelse(type == "NIL", molb_ped, paste0(grp, "_", prefix)),
         skim_prefix     = ifelse(prefix %in% skim_seq, prefix, NA_character_),
         pedigree_agrees = ifelse(type == "NIL", molb_ped == skim_ped, NA)) %>%
  left_join(brb_by_ped, by = c("molb_ped" = "pedigree")) %>%
  transmute(type, label, molb_prefix = prefix, skim_prefix, brbseq_prefix,
            molb_ped, skim_ped, pedigree_agrees,
            in_skim = !is.na(skim_prefix), in_brbseq = !is.na(brbseq_prefix)) %>%
  arrange(type, molb_prefix)
write_csv(file2, file.path(params$outdir, "molbreeding_3way_correspondence.csv"))
kable(file2, caption = "MolBreeding samples across all three experiments (MolB id · Skim id · BRB-seq id)")
```

All `r sum(file2$type=="NIL")` MolBreeding NILs share the **same prefix and pedigree** with
Skim (`pedigree_agrees = TRUE`) — confirming Skim and MolBreeding are one id space. **`r sum(file2$in_brbseq)`
of them were also sequenced in BRB-seq** (linked by pedigree; see `brbseq_prefix`). The 2
`B73-bulk` checks (`PN4_SID326`, `PN6_SID576`) **also carry a Skim id by shared prefix**
(`in_skim = TRUE` — they are among skim's B73 checks); having no pedigree, they have no
BRB-seq link.

## Checks to resolve {#checks-to-resolve}

NILs are matched by pedigree; **checks have no pedigree**, so only **Skim ↔ MolBreeding**
(one shared id space) can match them automatically — **BRB-seq cannot**. Here is exactly
what exists, and what still needs a key.

```{r checks}
checks <- bind_rows(
  tibble(experiment = "Skim",        group = skim_chk$grp, prefix = skim_chk$skim_prefix),
  tibble(experiment = "BRB-seq",     group = brb_chk$grp,  prefix = brb_chk$brb_prefix),
  tibble(experiment = "MolBreeding", group = "B73",
         prefix = file2$molb_prefix[file2$type != "NIL"]))
```

**How many checks each experiment sequenced** (rows are what need a cross-experiment id):

```{r checks-counts}
checks %>%
  count(experiment, group, name = "n_sequenced") %>%
  pivot_wider(names_from = experiment, values_from = n_sequenced, values_fill = 0) %>%
  kable(caption = "Sequenced check samples per experiment × group")
```

**The sequenced check samples** (none have a pedigree, so none can be matched by the NIL key):

```{r checks-table}
checks_tab <- checks %>% transmute(experiment, group, prefix) %>% arrange(group, experiment, prefix)
write_csv(checks_tab, file.path(params$outdir, "checks_to_resolve.csv"))
kable(checks_tab, caption = "Sequenced check samples (also written to checks_to_resolve.csv)")
```

Matching status:

- **Skim ↔ MolBreeding is already resolved** by shared id — the 2 MolBreeding `B73-bulk`
  (`PN4_SID326`, `PN6_SID576`) are the same samples as two of skim's B73 checks.
- **BRB-seq checks need a key.** They have their own id space and no pedigree, so they
  cannot be matched to the skim/MolBreeding checks automatically.

> **For Hannah:** is there a physical-sample key (plant / DNA aliquot / plate provenance)
> that maps the **BRB-seq** check wells to the Skim/MolBreeding checks? (Skim↔MolBreeding are
> already linked by shared id.) If B73 checks are interchangeable controls (each should read
> ≈ pure recurrent, ~0 donor), we may only need to treat them as a **control class per
> experiment** rather than match them 1:1. Please confirm the intended handling.

## Summary

```{r summary}
tibble(
  table = c("Skim ↔ BRB-seq", "MolBreeding 3-way"),
  key   = c("pedigree (.B-dropped)", "shared PN#_SID# (Skim); pedigree (BRB)"),
  NILs  = c(sum(nil1$in_skim & nil1$in_brbseq), sum(file2$type=="NIL")),
  note  = c("comparable on both platforms",
            sprintf("16 MolB samples; %d also in BRB-seq", sum(file2$in_brbseq))),
  file = file.path(params$outdir, c("skim_brbseq_correspondence.csv","molbreeding_3way_correspondence.csv"))
) %>% kable(caption = "Deliverables")
```

- **Skim ↔ BRB-seq:** `r sum(nil1$in_skim & nil1$in_brbseq)` comparable NILs (by pedigree).
- **MolBreeding 3-way:** all `r sum(file2$type=="NIL")` MolB NILs map to Skim (shared id);
  `r sum(file2$in_brbseq)` are also in BRB-seq (by pedigree).
- **Checks:** Skim↔MolBreeding matched by shared id; **BRB-seq checks** still need a key
  (see above).
