---
title: "Skim-RTI vs BrB-RTI — ancestry concordance on pedigree-matched NILs"
author: "Fausto Rodríguez-Zapata"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    df-print: paged
    embed-resources: true
execute:
  warning: false
  message: false
params:
  molb_3way:  "results/sample_correspondence/molbreeding_3way_correspondence.csv"        # MolB·Skim·BRB pedigree map; painting/confusion/Jaccard scope = NILs with in_brbseq==TRUE
  skim_calls: "data/rtiger_50K/calls_taxa_r5.csv"                                       # Skim-RTI, per-taxon r=5
  brb_meta:   "data/BRBseq_metadata.csv"                                                # BRB sample_id -> pedigree (coverage section)
  skim_ped:   "data/skim_sample_pedigree.csv"                                           # skim sample -> pedigree (.B dropped; coverage section)
  corr:       "results/sample_correspondence/skim_brbseq_correspondence.csv"            # curated NIL pairs (pedigree-matched, seq-restricted)
  mb_counts:  "results/sim_calibration/gbts_calls/SNP_wsfilt/counts"                    # wideseq grid (9,157)
  brb_cov:    "results/sim_calibration/brbseq_coverage.csv"                             # BRB-seq per-sample mean coverage (all 379)
  skim_cov:   "results/sim_calibration/real_skim_lambda.csv"                            # SNP50K skim per-sample mean coverage (all 1,393)
  brb_sweep:  "results/sim_calibration/brbseq_taxa_r7_wideseq/calls_common_schema.csv"   # BrB-RTI r=7 wideseq-filtered, per-species (sweep samples + B73 subset by name)
  skim_sweep: "results/sim_calibration/skim_sweep_calls/calls_common_schema.csv"        # Skim-RTI r=5 calls for the sweep samples + B73
  members:    "results/sim_calibration/coverage_sweep_members.csv"                      # the two 10-NIL coverage sweeps
  brb_taxa:   "results/sim_calibration/brbseq_taxa_r7_wideseq/calls_common_schema.csv"   # BrB-RTI r=7, wideseq-filtered, per teosinte species (full 379)
  nil_seg:    "results/nil_segments.csv"                                                # n=1500 BC2S3 simulation introgression segments (NULL reference)
---

**Repo:** [sawers-rellan-labs/zealtiger](https://github.com/sawers-rellan-labs/zealtiger)

Skim-RTI vs BrB-RTI ancestry concordance, matched **by NIL pedigree string** — not by
the `PN#_SID#` plate index, which is mislabeled in the BRB-seq run
(`docs/brbseq-sample-mislabeling.md`, `docs/brbseq_sample_identity.html`). The BRB-seq
plate id does not track the NIL; the **pedigree string is the cross-experiment
identifier**. The only formatting difference between the two metadata files is a
trailing **`.B`** the MolBreeding/skim registry appends and the BRB metadata omits, so
we **drop `.B`** before joining.

- BrB-RTI sample identity: `BRBseq_metadata.csv` (`sample_id` → `genotype`, drop `.B`).
- Skim-RTI sample identity: the BZea registry (`sample_metatada.csv` → staged as
  `data/skim_sample_pedigree.csv`, already `.B`-dropped).

The pedigree-matched **pairs are sourced from the curated correspondence**
(`results/sample_correspondence/skim_brbseq_correspondence.csv`, built by
`sample_correspondence.qmd`); this doc keeps the pairs whose ancestry calls are available
here. (The coverage section below still pairs the full dataset directly via the registries.)

The per-NIL painting / confusion / Jaccard cover the **4 NILs that have both
MolBreeding-GBTS and BRB-seq** (pedigree-matched: `Zv.0080`, `Zx.0340`, `Zx.0540`,
`Zd.0030`) — the lines with the most cross-platform support. The coverage and
introgression-size sections use the full skim∩BRB set. (The earlier 8 `PN4_SID#` samples
were a cross-id-space mislabeling artifact, not genuine MolBreeding∩BRB NILs — see
`docs/brbseq-sample-mislabeling.md`.)

## Setup

```{r setup}
library(tidyverse); library(data.table); library(GenomicRanges); library(ggpubr)
pal <- c(REF = "gold", HET = "springgreen4", ALT = "purple4")
chop <- function(x) sub("\\.B$", "", x)

brb  <- read_csv(params$brb_taxa,   show_col_types = FALSE)   # BrB-RTI, per-species fit (carries the real BRB sample names)
skim <- read_csv(params$skim_calls, show_col_types = FALSE)
brbmeta <- read_csv(params$brb_meta, show_col_types = FALSE) %>%
  transmute(sid = sample_id, nil = chop(genotype))
skimped <- read_csv(params$skim_ped, show_col_types = FALSE)   # sample, pedigree (.B-dropped)
corr <- read_csv(params$corr, show_col_types = FALSE) %>% filter(type == "NIL")  # full skim<->BRB map (intro-size section)

# Per-NIL painting / confusion / Jaccard scope = the MolBreeding-GBTS NILs that ALSO have
# BRB-seq, matched BY PEDIGREE (molbreeding_3way_correspondence.csv, in_brbseq == TRUE).
# Their real BRB libraries are PN3_SID265 / PN2_SID137 / PN1_SID10 / PN2_SID169 — NOT the
# PN4 plate names, which are a cross-id-space mislabeling artifact
# (docs/brbseq-sample-mislabeling.md). BrB-RTI calls come from the per-species fit (`brb`).
matched <- read_csv(params$molb_3way, show_col_types = FALSE) %>%
  filter(type == "NIL", in_brbseq) %>%
  transmute(nil = molb_ped, skim_name = skim_prefix, brb_name = brbseq_prefix)

# common wideseq grid + state-on-grid
cf   <- read_tsv(list.files(params$mb_counts, full.names = TRUE)[1],
                 col_names = c("chr_label","bp","rb","rc","ab","ac"), show_col_types = FALSE)
grid <- tibble(chr = as.integer(sub("chr","",cf$chr_label)), bp = cf$bp)
gg   <- GRanges(grid$chr, IRanges(grid$bp, width = 1))
state_on_grid <- function(seg, nm) {
  d <- seg[seg$name == nm, ]
  d$state[findOverlaps(gg, GRanges(d$chr, IRanges(d$start_bp, d$end_bp)), select = "first")]
}

cat(sprintf("Per-NIL scope — MolBreeding NILs that also have BRB-seq (pedigree-matched): %d\n  %s\n",
            nrow(matched), paste(matched$nil, collapse = ", ")))
```

## Coverage by platform — full dataset

Mean per-sample sequencing coverage on each platform, paired **by NIL pedigree**
across the **whole** dataset (every BRB-seq sample joined to its same-pedigree skim
sample). SNP50K-skim λ from `real_skim_lambda.csv` (mean depth over the 50K grid);
BRB-seq λ from `brbseq_coverage.csv` (mean depth over the brbseq SNP sites). One
point = one NIL. The mean of each platform is drawn as a dotted reference line.

```{r coverage-data}
brb_cov <- read_csv(params$brb_cov, show_col_types = FALSE) %>%
  mutate(sid = as.integer(sub(".*_SID", "", name)),
         nil = brbmeta$nil[match(sid, brbmeta$sid)]) %>%
  filter(!is.na(nil)) %>% transmute(nil, brb_name = name, brb_cov = lambda)
skim_cov <- read_csv(params$skim_cov, show_col_types = FALSE) %>%
  mutate(nil = skimped$pedigree[match(name, skimped$sample)]) %>%
  filter(!is.na(nil)) %>%
  group_by(nil) %>% summarise(skim_name = dplyr::first(name), skim_cov = dplyr::first(lambda), .groups = "drop")
cov <- inner_join(brb_cov, skim_cov, by = "nil") %>%
  filter(skim_name %in% skim$name)   # keep NILs that have a Skim-RTI call (so a pick is always paintable)
ms <- mean(cov$skim_cov); mb <- mean(cov$brb_cov)
cat(sprintf("%d pedigree-matched NILs | mean coverage: SNP50K skim %.3f× | BRB-seq %.3f×\n",
            nrow(cov), ms, mb))
```

### Coverage-matched series — grid the varying platform across its central 95%

To sweep one platform's coverage while controlling the other, we **grid the varying
platform across its central 95% (P2.5–P97.5)** on 10 evenly-spaced steps; at each step we
take the NIL **nearest the intersection of that grid line and the held platform's
mean**, measuring distance in the plot's *screen* space (each coverage axis scaled by
its plotted span, ≈ 10:1 skim:BRB). This pins picks to both the evenly-spaced grid and
the held-mean line (feasible because the two coverages are uncorrelated). The screen
weight was calibrated against hand-clicked picks on the skim sweep (reproduces 9/10).

```{r grid-sample}
# At each of the n grid steps, pick the NIL nearest the INTERSECTION of (grid_target on the
# varying axis, hold_mean on the held axis), with distance measured in *screen* space — each
# coverage axis scaled by its plotted span (weights `W`; BRB spans ~10× skim's range). This
# lands picks on both the evenly-spaced grid and the held-mean line. Calibrated to reproduce
# hand-clicked picks on the skim sweep (9/10); generalises to the BRB sweep by swapping which
# axis carries the grid (the weights make it symmetric).
W <- c(skim_cov = 1, brb_cov = 0.1)   # per-unit screen weight (skim:BRB ≈ 10:1)
grid_series <- function(df, vary, hold, hold_mean, n = 10, q = c(0.025, 0.975)) {
  v <- df[[vary]]; h <- df[[hold]]; wv <- W[[vary]]; wh <- W[[hold]]
  gr <- seq(quantile(v, q[1]), quantile(v, q[2]), length.out = n)
  chosen <- integer(0)
  for (g in gr) {
    free <- setdiff(seq_along(v), chosen)
    D <- (wv * (v[free] - g))^2 + (wh * (h[free] - hold_mean))^2   # screen-space dist to (g, hold_mean)
    chosen <- c(chosen, free[which.min(D)])
  }
  df[chosen, , drop = FALSE] %>% mutate(grid_target = gr)
}
vary_brb  <- grid_series(cov, "brb_cov",  "skim_cov", ms)   # sweep BRB-seq, hold skim ≈ mean
vary_skim <- grid_series(cov, "skim_cov", "brb_cov", mb)    # sweep skim,    hold BRB  ≈ mean
cat(sprintf("BRB-seq central 95%%: [%.2f, %.2f]× | skim central 95%%: [%.3f, %.3f]×\n",
            quantile(cov$brb_cov,.025), quantile(cov$brb_cov,.975),
            quantile(cov$skim_cov,.025), quantile(cov$skim_cov,.975)))
```

### Scatter — mean lines + the two gridded series

```{r coverage-scatter, fig.width=7.5, fig.height=6}
cov2 <- cov %>% mutate(series = case_when(
  nil %in% vary_brb$nil & nil %in% vary_skim$nil ~ "both series",
  nil %in% vary_brb$nil  ~ "vary BRB (skim≈mean)",
  nil %in% vary_skim$nil ~ "vary skim (BRB≈mean)",
  TRUE ~ "other"))
ggplot(cov2, aes(skim_cov, brb_cov)) +
  geom_vline(xintercept = ms, linetype = "dotted") +
  geom_hline(yintercept = mb, linetype = "dotted") +
  geom_point(aes(colour = series, size = series == "other"), alpha = 0.8) +
  scale_size_manual(values = c(`TRUE` = 1.3, `FALSE` = 2.7), guide = "none") +
  scale_colour_manual(values = c(other = "grey75", `vary BRB (skim≈mean)` = "firebrick",
                                 `vary skim (BRB≈mean)` = "royalblue3", `both series` = "purple3"),
                      name = NULL) +
  labs(x = "SNP50K skim mean coverage (×)", y = "BRB-seq mean coverage (×)",
       title = "Per-NIL coverage — SNP50K skim vs BRB-seq",
       subtitle = sprintf("dotted = platform mean (skim %.2f×, BRB %.2f×) | 10-point grid across each varying platform's central 95%%",
                          ms, mb)) +
  theme_bw()
```

### The two coverage-matched series

**Vary BRB-seq** across its central 95%, **skim held ≈ mean** — sorted by BRB-seq λ:

```{r series-brb}
vary_brb %>% arrange(brb_cov) %>%
  transmute(NIL = nil, brb_cov = round(brb_cov,3), skim_cov = round(skim_cov,3),
            grid_target = round(grid_target,2), brb_sample = brb_name, skim_sample = skim_name) %>%
  knitr::kable(caption = "Skim held ≈ mean; BRB-seq λ stepped across its central 95% (P2.5–P97.5).")
```

**Vary skim** across its central 95%, **BRB-seq held ≈ mean** — sorted by skim λ:

```{r series-skim}
vary_skim %>% arrange(skim_cov) %>%
  transmute(NIL = nil, skim_cov = round(skim_cov,3), brb_cov = round(brb_cov,3),
            grid_target = round(grid_target,3), skim_sample = skim_name, brb_sample = brb_name) %>%
  knitr::kable(caption = "BRB-seq held ≈ mean; skim λ stepped across its central 95% (P2.5–P97.5).")
```

**Read:** BRB-seq (3′ RNA-seq) mean depth (~2.8×) sits well above the ~0.4× SNP50K
skim, and the two platforms' coverages are uncorrelated across NILs (library-driven).
Each series steps one platform evenly across its central-95% coverage while holding
the other ≈ its mean — a coverage-controlled sweep for asking how the ancestry calls
move with the varying platform's depth, with the other platform's coverage held out
as a confounder.

## Coverage-sweep chromosome painting

Ancestry calls for the two 10-NIL coverage sweeps, painted as the coverage of the
*varying* platform climbs (rows ordered low→high; the row label prefixes the varying
λ). Two tracks per NIL — **top = Skim-RTI (r=5), bottom = BrB-RTI (r=7, ws)** — and a
**B73 control** row pinned at top (forced, not pedigree-matched: skim `PN10_SID893`,
BRB `PN3_SID213` — clean reps; the old `PN1_SID15` is an introgressed impostor,
`docs/b73_checks_qc.html`). Each B73 control comes from its platform's **B73 check group**,
fit separately from the teosinte donors.

**BrB-RTI = wideseq-filtered, per-species, r=7.** BRB-seq 3′ markers are dense and noisy,
so on the full joint-called SNP set RTIGER needed an extreme rigidity (r=109). Restricting
the BRB allele counts to the **wideseq teosinte-informative positions** (21,665 of 66,983
SNPs; `data/brbseq_wideseq/brb_wideseq.vcf.gz`, GATK SelectVariants on the Chen2022 panel)
strips the off-target noise and drops the **KS-calibrated rigidity to r=7** — chosen by
minimizing `D = sup_x|F_called − F_sim|` of the donor-block size distribution vs the n=1500
BC2S3 simulation (`D=0.103`, vs 0.112 at the old r=109; median donor block 11.3 Mb ≈ the
sim's 10.74). Curve in `results/sim_calibration/brbseq_ks_wideseq/brbseq_ks_refined.csv`.
BrB-RTI is fit **per teosinte species** (*Z. mays* ssp. *parviglumis* `Zv`, ssp. *mexicana*
`Zx`, ssp. *huehuetenangensis* `Zh`, *Z. diploperennis* `Zd`, *Z. luxurians* `Zl`), B73 and
Purple as separate checks (`brbseq_taxa_r7_wideseq`). The B73 check-group floor stays low
(median ~0, mean ~0.05); the control `PN3_SID213` reads ≈ 0 donor.

```{r sweep-paint-setup}
brb_sw  <- read_csv(params$brb_sweep,  show_col_types = FALSE)
skim_sw <- read_csv(params$skim_sweep, show_col_types = FALSE)
mem     <- read_csv(params$members,    show_col_types = FALSE)
BRB_B73 <- "PN3_SID213"; SKIM_B73 <- "PN10_SID893"   # clean B73 reps (PN1_SID15 is an introgressed impostor; see docs/b73_checks_qc.html)

lane2 <- function(seg, nm, method, rowlab) {
  d <- seg %>% filter(name == nm)
  if (!nrow(d)) return(NULL)
  tibble(row = rowlab, chr = d$chr, start_bp = d$start_bp, end_bp = d$end_bp,
         state = factor(d$state, 0:2, c("REF","HET","ALT")), method = method)
}
sweep_paint <- function(which) {
  m <- mem %>% filter(sweep == which) %>% arrange(vary_cov) %>%
    mutate(rowlab = sprintf("%.2f×<br>%s", vary_cov, gsub("_", "<br>", nil)))
  nil_rows <- pmap_dfr(list(m$skim_name, m$brb_name, m$rowlab), function(sk, br, rl)
    bind_rows(lane2(skim_sw, sk, "Skim-RTI", rl), lane2(brb_sw, br, "BrB-RTI", rl)))
  b73 <- bind_rows(lane2(skim_sw, SKIM_B73, "Skim-RTI", "**B73 control**"),
                   lane2(brb_sw,  BRB_B73,  "BrB-RTI",  "**B73 control**"))
  # B73 control pinned at the TOP; coverage rows keep their current descending order below
  bind_rows(b73, nil_rows) %>%
    mutate(chr = factor(chr, levels = 1:10, labels = paste0("chr", 1:10)),
           row = factor(row, levels = c("**B73 control**", rev(m$rowlab))),
           method = factor(method, c("Skim-RTI","BrB-RTI")),
           ylo = if_else(method == "Skim-RTI", 0.52, 0.00),
           yhi = if_else(method == "Skim-RTI", 1.00, 0.48))
}
paint_plot <- function(df, title, sub) {
  ggplot(df, aes(xmin = start_bp/1e6, xmax = end_bp/1e6, ymin = ylo, ymax = yhi, fill = state)) +
    geom_rect() +
    facet_grid(row ~ chr, scales = "free_x", space = "free_x", switch = "y") +
    scale_fill_manual(values = pal, drop = FALSE) +
    scale_x_continuous(labels = scales::unit_format(unit = "", scale = 1)) +
    scale_y_continuous(breaks = c(0.24, 0.76), labels = c("BrB-RTI","Skim-RTI"),
                       expand = expansion(mult = 0.02), position = "right") +
    labs(title = title, subtitle = sub, x = "Position (Mb)", fill = "Genotype") +
    ggpubr::theme_classic2(base_size = 9) +
    theme(legend.position = "bottom", strip.background = element_blank(),
          strip.text.x = element_text(face = "bold"),
          strip.text.y.left = ggtext::element_markdown(angle = 0, hjust = 1, size = 9, lineheight = 0.8),
          panel.spacing.x = unit(0.05, "lines"), panel.spacing.y = unit(0.18, "lines"),
          axis.text.y.right = element_text(size = 9), axis.ticks.y = element_blank(),
          axis.title.y = element_blank(), axis.line.y = element_blank(),
          axis.text.x = element_text(size = 6))
}
```

### Vary BRB-seq (skim ≈ mean)

```{r sweep-vary-brb, fig.width=12, fig.height=12}
paint_plot(sweep_paint("vary_brb"),
           "Coverage sweep — vary BRB-seq depth, skim held ≈ mean",
           "rows: B73 control then 10 NILs by increasing BRB-seq λ | top Skim-RTI / bottom BrB-RTI | gold REF, green HET, purple ALT")
```

### Vary skim (BRB-seq ≈ mean)

```{r sweep-vary-skim, fig.width=12, fig.height=12}
paint_plot(sweep_paint("vary_skim"),
           "Coverage sweep — vary skim depth, BRB-seq held ≈ mean",
           "rows: B73 control then 10 NILs by increasing skim λ | top Skim-RTI / bottom BrB-RTI")
```

**Read:** the **B73 control** anchors both panels — Skim-RTI paints it ~all-REF
(gold), while BrB-RTI speckles it with spurious het/donor (its RNA false-donor
floor). Down the sweeps, watch how each platform's calls fill in (or fragment) as its
own depth rises and as the other platform's depth stays fixed.

## Pedigree-matched pairs

```{r pairs-table}
jac_one <- function(s_nil) {
  p <- matched[matched$nil == s_nil, ]
  b <- state_on_grid(brb,  p$brb_name); k <- state_on_grid(skim, p$skim_name)
  ok <- !is.na(b) & !is.na(k); bd <- b[ok] > 0; kd <- k[ok] > 0
  u <- sum(bd | kd)
  tibble(donor_J = if (u == 0) NA_real_ else sum(bd & kd)/u,
         concord = mean(b[ok] == k[ok]),
         brb_donor = mean(bd), skim_donor = mean(kd))
}
pair_tab <- matched %>% mutate(m = map(nil, jac_one)) %>% unnest(m) %>%
  transmute(NIL = nil, brb_label = brb_name, skim_sample = skim_name,
            donor_J = round(donor_J, 2), concord = round(concord, 3),
            brb_donor = round(brb_donor, 3), skim_donor = round(skim_donor, 3)) %>%
  arrange(desc(donor_J))
knitr::kable(pair_tab,
             caption = "Skim-RTI vs BrB-RTI per pedigree-matched NIL: donor Jaccard, marker concordance, donor fraction. (PN#_SID# plate labels shown for provenance only.)")
```

**Read:** high `donor_J` = the two platforms agree on *where* the teosinte is for that
NIL. A low value where the pedigree matches but the genotype doesn't (watch `Zx.0370`)
flags a sample whose BRB-seq content disagrees with its stated pedigree — a residual
identity problem beyond the `.B` fix.

## Confusion matrix — Skim-RTI vs BrB-RTI (pooled, matched NILs)

```{r confusion}
paired <- map_dfr(matched$nil, function(s_nil) {
  p <- matched[matched$nil == s_nil, ]
  b <- state_on_grid(brb, p$brb_name); k <- state_on_grid(skim, p$skim_name)
  ok <- !is.na(b) & !is.na(k)
  tibble(nil = s_nil, skim = k[ok], brb = b[ok])
})
cm <- table(skim = factor(paired$skim, 0:2), brb = factor(paired$brb, 0:2))
prop <- sweep(cm, 1, rowSums(cm), "/")
disp <- matrix(sprintf("%d (%.1f%%)", cm, 100*prop), nrow = 3,
               dimnames = list(`skim \\ brb` = c("0 recurrent","1 het","2 donor"),
                               c("0 recurrent","1 het","2 donor")))
knitr::kable(disp, align = "r",
             caption = sprintf("Skim-RTI (rows) vs BrB-RTI (cols): count (row %%). Overall concordance %.3f.",
                               sum(diag(cm))/sum(cm)))
dp <- paired$skim %in% c(1,2)
cat(sprintf("Skim donor-present markers: %d | BrB also donor-present: %.1f%%\n",
            sum(dp), 100*mean(paired$brb[dp] %in% c(1,2))))
```

## Introgression-size distribution

Donor-block sizes (runs of donor state — HET or ALT — collapsed per chromosome) pooled
across the **full set of NILs called on *both* platforms** (pedigree-matched via the
correspondence; BrB-RTI from the per-species fit, all teosinte species). Grey curve is
the **Null** n=1500 BC2S3 simulation — the modeled expectation under no selection.

```{r intro-sizes, fig.width=8, fig.height=5}
brb_taxa <- read_csv(params$brb_taxa, show_col_types = FALSE)   # full per-species BrB-RTI (379)
nil_seg  <- read_csv(params$nil_seg,  show_col_types = FALSE)   # n=1500 sim (NULL)

# FULL set of NILs with calls on BOTH platforms (pedigree-matched, ";"-split replicates)
brb_in  <- corr %>% separate_rows(brbseq_prefix, sep = ";") %>%
  transmute(name = brbseq_prefix, nil = pedigree) %>% filter(name %in% brb_taxa$name)
skim_in <- corr %>% separate_rows(skim_prefix,  sep = ";") %>%
  transmute(name = skim_prefix,  nil = pedigree) %>% filter(name %in% skim$name)
both_nils <- intersect(brb_in$nil, skim_in$nil)

# introgression block = maximal run of donor (state>0) within a (sample, chromosome)
donor_blocks <- function(seg) seg %>% arrange(name, chr, start_bp) %>% group_by(name, chr) %>%
  mutate(run = cumsum(c(TRUE, diff(state > 0) != 0)), donor = state > 0) %>% filter(donor) %>%
  group_by(name, chr, run) %>% summarise(size_mb = (max(end_bp) - min(start_bp))/1e6, .groups = "drop")

skim_b <- donor_blocks(skim     %>% filter(name %in% skim_in$name[skim_in$nil %in% both_nils])) %>%
  mutate(src = "Skim-RTI (r=5)")
brb_b  <- donor_blocks(brb_taxa %>% filter(name %in% brb_in$name[brb_in$nil  %in% both_nils])) %>%
  mutate(src = "BrB-RTI (r=7, ws)")
null_b <- nil_seg %>% transmute(size_mb = mb)
cat(sprintf("NILs called on both platforms: %d | Skim samples %d (%d blocks) | BrB samples %d (%d blocks)\n",
            length(both_nils), n_distinct(skim_in$name[skim_in$nil %in% both_nils]), nrow(skim_b),
            n_distinct(brb_in$name[brb_in$nil %in% both_nils]), nrow(brb_b)))

ggplot(bind_rows(skim_b, brb_b), aes(x = size_mb, fill = src)) +
  geom_histogram(aes(y = after_stat(density)), position = "identity", alpha = 0.5, bins = 30) +
  geom_freqpoly(data = null_b, aes(x = size_mb, y = after_stat(density)),
                bins = 30, colour = "grey20", linewidth = 0.9, inherit.aes = FALSE) +
  scale_fill_manual(values = c("Skim-RTI (r=5)" = "purple4", "BrB-RTI (r=7, ws)" = "orange3")) +
  scale_x_log10() +
  labs(x = "introgression size (Mb, log scale)", y = "density", fill = NULL,
       title = "Introgression-size distribution — Skim-RTI vs BrB-RTI",
       subtitle = sprintf("%d NILs called on both platforms (pedigree-matched, full set) | grey = Null (n=1500 sim)",
                          length(both_nils))) +
  theme_bw()
```

## Chromosome painting — Skim-RTI vs BrB-RTI per NIL

Two tracks per NIL (**top = Skim-RTI r=5, bottom = BrB-RTI r=7 wideseq**), grouped and labelled
by the `.B`-dropped **pedigree** (wrapped at the left); the method label is at the right.
NILs ordered by donor Jaccard (best agreement first).

```{r painting, fig.width=12, fig.height=11}
nil_ord <- pair_tab$NIL
lane <- function(seg, nm, method, s_nil) {
  d <- seg[seg$name == nm, ]
  tibble(nil = s_nil, chr = d$chr, start_bp = d$start_bp, end_bp = d$end_bp,
         state = factor(d$state, 0:2, c("REF","HET","ALT")), method = method)
}
paint <- map_dfr(matched$nil, function(s_nil) {
  p <- matched[matched$nil == s_nil, ]
  bind_rows(lane(skim, p$skim_name, "Skim-RTI", s_nil),
            lane(brb,  p$brb_name,  "BrB-RTI",  s_nil))
}) %>%
  mutate(chr  = factor(chr, levels = 1:10, labels = paste0("chr", 1:10)),
         nil  = factor(gsub("_", "<br>", nil), levels = gsub("_", "<br>", nil_ord)),
         method = factor(method, c("Skim-RTI","BrB-RTI")),
         ylo = if_else(method == "Skim-RTI", 0.52, 0.00),
         yhi = if_else(method == "Skim-RTI", 1.00, 0.48))

ggplot(paint, aes(xmin = start_bp/1e6, xmax = end_bp/1e6, ymin = ylo, ymax = yhi, fill = state)) +
  geom_rect() +
  facet_grid(nil ~ chr, scales = "free_x", space = "free_x", switch = "y") +
  scale_fill_manual(values = pal, drop = FALSE) +
  scale_x_continuous(labels = scales::unit_format(unit = "", scale = 1)) +
  scale_y_continuous(breaks = c(0.24, 0.76), labels = c("BrB-RTI","Skim-RTI"),
                     expand = expansion(mult = 0.02), position = "right") +
  labs(title = "Skim-RTI vs BrB-RTI — pedigree-matched NILs",
       subtitle = "per NIL: top = Skim-RTI (r=5), bottom = BrB-RTI (r=7, ws) | gold REF, green HET, purple ALT",
       x = "Position (Mb)", fill = "Genotype") +
  ggpubr::theme_classic2(base_size = 9) +
  theme(legend.position = "bottom", strip.background = element_blank(),
        strip.text.x = element_text(face = "bold"),
        strip.text.y.left = ggtext::element_markdown(angle = 0, hjust = 1, size = 9, lineheight = 0.85),
        panel.spacing.x = unit(0.05, "lines"), panel.spacing.y = unit(0.15, "lines"),
        axis.text.y.right = element_text(size = 9), axis.ticks.y = element_blank(),
        axis.title.y = element_blank(), axis.line.y = element_blank(),
        axis.text.x = element_text(size = 6))
```

## Findings

- The per-NIL painting, confusion, and Jaccard cover the **4 MolBreeding-GBTS NILs that
  also have BRB-seq** (`Zv.0080`, `Zx.0340`, `Zx.0540`, `Zd.0030`; matched by pedigree) —
  the lines with the most cross-platform support. Coverage and introgression-size use the
  full skim∩BRB set.
- Where the pedigrees genuinely correspond, Skim-RTI and BrB-RTI agree on the donor
  landscape (high donor Jaccard); BrB-RTI is the sparser, noisier track (low coverage RNA).
- Any matched NIL with low Jaccard (e.g. `Zx.0370`) is a sample whose BRB-seq genotype
  disagrees with its stated pedigree — a residual identity issue the `.B` fix does not
  resolve, to be confirmed by the Hazel rerun (`docs/brbseq-sample-mislabeling.md`).
```
